---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 8,
  fig.height = 5,
  warning   = FALSE,
  message   = FALSE
)
```

```{r pkgs}
library(fastblm)
library(spatintegrate)
library(goebel2026)
library(Matrix)
library(sf)
library(ggplot2)
library(patchwork)

set.seed(42)
```

## Preview

The model fit looks like this, and in this demo we'll build the pieces and show how to tune parameters:

```{r eval = FALSE}
fit_aug <- fastblm::fit_fastblm(
  y      = y_alb,
  A      = A_aug,
  Q      = Q_aug,
  phi    = phi_aug,
  solver = "cholesky"
)
```


## Data

### Projected sf observations / target


```{r projected}
# Ensures each dataset is the same projected-coordinates before computing area-intersections
soundings_proj <- spatintegrate::ensure_projected(goebel2026::soundings_augmented)
target_proj    <- spatintegrate::ensure_projected(goebel2026::target_grid)

```

### A-matrix construction using spatintegrate

```{r Amatrix}

A <- as(
  spatintegrate::compute_overlap_fractions(soundings_proj, target_proj),
  "dgCMatrix"
)

p <- ncol(A)

```

### Observation-vector simulation

```{r observation-vector}
truth <- goebel2026::target_grid$mean_albedo

# Raw upscaled
# y_alb <- goebel2026::soundings_augmented$mean_albedo +
#   rnorm(length(goebel2026::soundings_augmented$mean_albedo),
#         0, noise_sd)

# Upscale with A matrix
noise_sd <- sd(goebel2026::target_grid$mean_albedo, na.rm = TRUE) / 20

y_alb <- A %*% truth +
  rnorm(length(goebel2026::soundings_augmented$mean_albedo),
        0, noise_sd)

```

## Covariates

```{r covariates}
water_sounding <- goebel2026::soundings_augmented$proportion_water
X_fixed <- cbind(intercept = 1, water = water_sounding)
q       <- ncol(X_fixed)

water_grid_p <- goebel2026::target_grid$proportion_water
water_grid_p[is.na(water_grid_p)] <- 0

X_grid <- cbind(intercept = 1, water = water_grid_p)
```



## Intrinsic SAR Prior (ρ = 1)

Setting ρ = 1 gives an intrinsic SAR prior: Q = (I − W)ᵀ(I − W) is positive
semi-definite (singular). K\_aug is still invertible because AᵀA regularises
the system in the observed directions. Because Q is singular its log-determinant
is undefined; we use `log_det_Q = 0` as a convention (improper prior) — it is a
constant that cancels out of the φ optimum.

```{r Q}

W <- goebel2026::make_W_matrix(goebel2026::target_grid)

rho_fixed <- 1.0
IminusW   <- Matrix::Diagonal(nrow(W)) - rho_fixed * W
Q_fixed   <- Matrix::forceSymmetric(Matrix::crossprod(IminusW))
Q_fixed   <- Matrix::drop0(Q_fixed)

cat(sprintf("Q fixed at rho=1: p=%d  nnz=%d\n", nrow(Q_fixed), nnzero(Q_fixed)))
```

## Tune φ via CV

```{r tune}
lambda_beta <- 0.01

A_aug <- as(cbind(A, X_fixed), "dgCMatrix")

Q_aug_fun <- function(theta) {
  Q_aug     <- Matrix::bdiag(Q_fixed, lambda_beta * Matrix::Diagonal(q))
  log_det_Q <- q * log(lambda_beta)
  list(Q = Q_aug, log_det_Q = log_det_Q)
}

tuned_cv_rho1 <- tune_cv(
  y          = y_alb,
  A          = A_aug,
  Q_fun      = Q_aug_fun,
  theta_init = numeric(0), # providing theta_init will 
 # X_cov      = X_fixed,
  k          = 5L,
  verbose    = TRUE
)

# ML alternative (faster, no cross-validation):
# tuned_ml_rho1 <- tune_ml(
#   y          = y_alb,
#   A          = A_aug,
#   Q_fun      = Q_aug_fun,
#   X_fixed    = NULL,
#   theta_init = numeric(0),
#   verbose    = TRUE
# )
```

## Fit

```{r fit}
phi_aug <- tuned_cv_rho1$phi
print(phi_aug)

Q_aug <- Matrix::bdiag(
  Q_fixed,
  lambda_beta / phi_aug * Matrix::Diagonal(q)
)

fit_aug <- fastblm::fit_fastblm(
  y      = y_alb,
  A      = A_aug,
  Q      = Q_aug,
  phi    = phi_aug,
  solver = "cholesky"
)

r_hat    <- fit_aug$posterior_mean[seq_len(p)]
beta_hat <- fit_aug$posterior_mean[p + seq_len(q)]

cat(sprintf("beta_hat: intercept=%.4f  water=%.4f\n", beta_hat[1], beta_hat[2]))

se_aug  <- fastblm::posterior_se(fit_aug)
se_r    <- se_aug[seq_len(p)]
se_beta <- se_aug[p + seq_len(q)]

cat(sprintf("se_beta:  intercept=%.4f  water=%.4f\n", se_beta[1], se_beta[2]))

# A_pred maps [r; beta] -> r + X_grid %*% beta, i.e. full grid-level prediction
A_pred  <- as(cbind(Matrix::Diagonal(p), X_grid), "dgCMatrix")
pred    <- as.vector(A_pred %*% fit_aug$posterior_mean)
se_pred <- fastblm::posterior_se(fit_aug, A_new = A_pred)
```

## RSR Constraint

```{r rsr}
C_spatial   <- as.matrix(t(X_fixed) %*% A)
C_aug       <- cbind(C_spatial, matrix(0, q, q))

fit_aug_rsr <- fastblm::constrain(fit_aug, C_aug)

r_hat_rsr    <- fit_aug_rsr$posterior_mean[seq_len(p)]
beta_hat_rsr <- fit_aug_rsr$posterior_mean[p + seq_len(q)]

cat(sprintf("RSR beta_hat: intercept=%.4f  water=%.4f\n",
            beta_hat_rsr[1], beta_hat_rsr[2]))
cat(sprintf("RSR ||C r||_inf = %.2e\n", max(abs(C_spatial %*% r_hat_rsr))))

# Sanity check on C Sigma C' conditioning
SigmaCt_rsr  <- fit_aug_rsr$constraint$SigmaCt
CSigmaCt_rsr <- as.matrix(C_aug) %*% as.matrix(SigmaCt_rsr)
cat(sprintf("C Sigma C' eigenvalues: %s\n",
            paste(formatC(eigen(CSigmaCt_rsr)$values, format = "e", digits = 3),
                  collapse = ", ")))

pred_rsr    <- as.vector(A_pred %*% fit_aug_rsr$posterior_mean)
se_pred_rsr <- fastblm::posterior_se(fit_aug_rsr, A_new = A_pred)
```

## Evaluation

```{r metrics}
# Restrict to cells covered by at least one sounding footprint -- unobserved
# cells are pure prior smoothing and shouldn't count toward predictive metrics.
observed_cells <- colSums(A) > 0

eval_metrics <- function(label, pred, se, truth) {
  keep  <- observed_cells & !is.na(truth)
  resid <- pred[keep] - truth[keep]
  rmse     <- sqrt(mean(resid^2))
  r2       <- 1 - sum(resid^2) / sum((truth[keep] - mean(truth[keep]))^2)
  coverage <- mean(truth[keep] >= pred[keep] - 1.96 * se[keep] &
                   truth[keep] <= pred[keep] + 1.96 * se[keep])
  cat(sprintf("%-22s  RMSE=%.4f  R2=%.4f  95%%cov=%.1f%%  (n=%d cells)\n",
              label, rmse, r2, 100 * coverage, sum(keep)))
  data.frame(Model = label, RMSE = rmse, R2 = r2, Coverage = coverage)
}

met_base <- eval_metrics("Unconstrained",   pred,     se_pred,     truth)
met_rsr  <- eval_metrics("RSR-constrained", pred_rsr, se_pred_rsr, truth)

knitr::kable(rbind(met_base, met_rsr), digits = 4, row.names = FALSE,
             col.names = c("Model", "RMSE", "R²", "95% Coverage"))
```

## Plots

```{r plot-setup, include=FALSE}
plot_sf <- target_proj
plot_sf$truth     <- truth
plot_sf$water     <- water_grid_p
plot_sf$r_hat     <- pred
plot_sf$r_hat_rsr <- pred_rsr
plot_sf$se        <- se_pred
plot_sf$se_rsr    <- se_pred_rsr
plot_sf$resid     <- pred     - truth
plot_sf$resid_rsr <- pred_rsr - truth
plot_sf$cov_base  <- !is.na(truth) &
                     truth >= pred     - 1.96 * se_pred &
                     truth <= pred     + 1.96 * se_pred
plot_sf$cov_rsr   <- !is.na(truth) &
                     truth >= pred_rsr - 1.96 * se_pred_rsr &
                     truth <= pred_rsr + 1.96 * se_pred_rsr

sf_plot <- function(data, fill, title,
                    lo = "#2c7bb6", mi = "#ffffbf", hi = "#d7191c",
                    midpt = NULL, lims = NULL) {
  if (is.null(midpt)) midpt <- mean(data[[fill]], na.rm = TRUE)
  ggplot(data) +
    geom_sf(aes(fill = .data[[fill]]), colour = NA) +
    scale_fill_gradient2(low = lo, mid = mi, high = hi,
                         midpoint = midpt, limits = lims,
                         na.value = "grey90", name = NULL) +
    labs(title = title) +
    theme_void(base_size = 10) +
    theme(plot.title = element_text(face = "bold", size = 10),
          legend.key.height = unit(0.5, "cm"))
}

lim_alb <- range(c(truth, r_hat, r_hat_rsr), na.rm = TRUE)
lim_se  <- range(c(se_pred, se_pred_rsr),    na.rm = TRUE)
lim_res <- max(abs(c(plot_sf$resid, plot_sf$resid_rsr)), na.rm = TRUE) * c(-1, 1)
```

### True field and water covariate

```{r plot-truth, fig.height=4}
sf_plot(plot_sf, "truth", "True albedo", lims = lim_alb, midpt = mean(lim_alb)) |
sf_plot(plot_sf, "water", "Water fraction",
        lo = "#f7fbff", mi = "#6baed6", hi = "#08306b", midpt = 0.25)
```

### Fitted fields

```{r plot-fits, fig.height=4}
(sf_plot(plot_sf, "r_hat",     "Posterior mean — unconstrained",
         lims = lim_alb, midpt = mean(lim_alb)) |
 sf_plot(plot_sf, "r_hat_rsr", "Posterior mean — RSR",
         lims = lim_alb, midpt = mean(lim_alb))) +
  plot_annotation(title = sprintf("phi = %.4f", phi_aug))
```

### Posterior SE

```{r plot-se, fig.height=4}
sf_plot(plot_sf, "se",    "SE — unconstrained",
        lo = "#fff7ec", mi = "#fc8d59", hi = "#7f0000",
        midpt = mean(lim_se), lims = lim_se) |
sf_plot(plot_sf, "se_rsr", "SE — RSR",
        lo = "#fff7ec", mi = "#fc8d59", hi = "#7f0000",
        midpt = mean(lim_se), lims = lim_se)
```

### Residuals

```{r plot-resid, fig.height=4}
sf_plot(plot_sf, "resid",     "Residual — unconstrained", lims = lim_res, midpt = 0) |
sf_plot(plot_sf, "resid_rsr", "Residual — RSR",           lims = lim_res, midpt = 0)
```

### Predicted vs. true

```{r scatter, fig.height=4}
keep  <- !is.na(truth)
sc_df <- data.frame(
  truth  = rep(truth[keep], 2),
  fitted = c(pred[keep], pred_rsr[keep]),
  model  = rep(c("Unconstrained", "RSR"), each = sum(keep))
)

ggplot(sc_df, aes(truth, fitted)) +
  geom_point(aes(colour = model), alpha = 0.3, size = 0.8) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  scale_colour_manual(values = c("Unconstrained" = "steelblue", "RSR" = "tomato"),
                      guide = "none") +
  facet_wrap(~ model) +
  labs(title = "Predicted vs. true albedo", x = "True", y = "Posterior mean") +
  theme_minimal(base_size = 13)
```

### 95% coverage

```{r coverage-map, fig.height=4}
cov_sf <- rbind(
  st_sf(covered = plot_sf$cov_base, model = "Unconstrained", geometry = st_geometry(plot_sf)),
  st_sf(covered = plot_sf$cov_rsr,  model = "RSR",           geometry = st_geometry(plot_sf))
)

ggplot(cov_sf) +
  geom_sf(aes(fill = covered), colour = NA) +
  scale_fill_manual(values  = c("TRUE" = "#d9f0d3", "FALSE" = "#e9a3c9"),
                    labels  = c("TRUE" = "Covered", "FALSE" = "Not covered"),
                    na.value = "grey90", name = NULL) +
  facet_wrap(~ model) +
  labs(title = "95% posterior credible interval coverage") +
  theme_void(base_size = 13)
```

## Session Info

```{r session}
sessionInfo()
```

