---
title: "Supplementary Materials: Bayesian Downscaling of Satellite-Derived Solar-Induced Chlorophyll Fluorescence"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    latex_engine: pdflatex
    keep_tex: false
header-includes:
  - \usepackage{booktabs}
  - \usepackage{float}
  - \usepackage{amsmath}
  - \usepackage{caption}
  - \captionsetup{font=small}
  - \floatplacement{figure}{H}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = FALSE,
  warning   = FALSE,
  message   = FALSE,
  fig.align = "center",
  fig.pos   = "H",
  out.width = "\\textwidth"
)

library(goebel2026)
library(fastblm)
library(Matrix)
library(ggplot2)
library(patchwork)
library(sf)
library(knitr)
# library(kableExtra) -- removed: requires multirow.sty not available on cluster

`%||%` <- function(x, y) if (!is.null(x)) x else y

load_result <- function(nm) {
  env <- new.env(parent = emptyenv())
  tryCatch(
    data(list = nm, package = "goebel2026", envir = env),
    warning = function(w) NULL, error = function(e) NULL
  )
  env[[nm]]
}

expand_to_grid <- function(vals, idx, n) {
  out <- rep(NA_real_, n)
  out[idx] <- vals
  out
}

d_shared   <- goebel2026::setup_shared
d_sif      <- goebel2026::setup_sif
target_sf  <- goebel2026::target_grid
target_idx <- which(target_sf$n_intersects > 0)
n_grid     <- nrow(target_sf)
```

\newpage

# Water Covariate and Spatial Confounding in SIF Downscaling {#sec:water}

## Motivation

SIF over open water is expected to be near zero, as photosynthesis does not
occur on water surfaces. This expectation is consistent with lower raw SIF
values in OCO-2 soundings whose footprints overlap with the Charles River
and Boston Harbour. Including water proportion as a covariate allows the
model to account for this mean difference explicitly.

However, including a spatially structured covariate alongside a spatial
random effect introduces spatial confounding (Hodges and Reich 2010): the
model may explain spatial variation through either the covariate coefficient
$\beta$ or the residual field $r^{(\ell)}$, making inference on $\beta$
unreliable. We document the problem and two solutions: hard restricted
spatial regression (RSR) and a soft penalty variant.

## Naive fit: unconstrained covariate

Without the RSR constraint, the spatial random effect and the water
covariate share explanatory power. Table \ref{tab:betas-naive} shows
$\hat\beta$ estimates under unconstrained models.

```{r betas-naive}
res_naive <- load_result("results_sif_rsr")
res_rho1  <- load_result("results_sif_rsr_rho1")
res_wz    <- load_result("results_sif_water_zero")

rows <- list(
  list(res_naive, "Water covariate, no RSR, $\\rho$ free"),
  list(res_rho1,  "Water covariate, no RSR, $\\rho=1$"),
  list(res_wz,    "Water zero constraint")
)

beta_df <- do.call(rbind, lapply(rows, function(x) {
  r <- x[[1]]; nm <- x[[2]]
  if (is.null(r) || is.null(r$beta_hat)) return(NULL)
  data.frame(Model     = nm,
             Intercept = round(r$beta_hat[1], 3),
             Water     = round(r$beta_hat[2], 3),
             rho       = round(r$rho_opt %||% 1, 3),
             phi       = round(r$phi, 3),
             stringsAsFactors = FALSE)
}))

if (!is.null(beta_df) && nrow(beta_df) > 0) {
  kable(beta_df, format = "latex", booktabs = TRUE, escape = FALSE,
        caption = paste0("Estimated covariate coefficients under naive (no RSR) models. ",
                         "The intercept estimates mean SIF over non-water pixels; ",
                         "the water coefficient estimates the offset for water pixels. ",
                         "Positive or near-zero water coefficients are inconsistent ",
                         "with the physical expectation that water SIF $\\approx 0$."),
        label = "betas-naive")
} else {
  cat("Naive SIF results not available.")
}
```

In the unconstrained model, the spatial random effect absorbs covariate
variation. The water coefficient is not reliably interpretable because the
model can explain the same data either through $\beta$ or through the
spatial field, depending on the smoothing parameter. When $\phi$ is large
(strong smoothing), more variation is attributed to the spatial field and
the covariate coefficient is compressed toward zero; when $\phi$ is small,
the covariate absorbs more variation. This instability is the signature of
spatial confounding.

## RSR: enforcing orthogonality

The RSR constraint
$$
X^{(o)\top} A r^{(\ell)} = 0
$$
ensures that the spatial component cannot explain variation collinear with
the covariate. Under this constraint, $\hat\beta$ equals the coefficient
that would be obtained if the model were fitted without a spatial random
effect at all, and is therefore interpretable as the pure water-effect
estimate. Table \ref{tab:betas-rsr} shows $\hat\beta$ under RSR-constrained
models.

```{r betas-rsr}
res_rsr_rho1  <- load_result("results_sif_rsr_rho1")
res_canonical <- load_result("results_sif_canonical")
res_softrsr   <- load_result("results_softrsr_albedo")

rows_rsr <- list(
  list(res_rsr_rho1,  "Hard RSR, $\\rho=1$"),
  list(res_canonical, "Hard RSR + $R^{-1}$, $\\rho=1$ (canonical)"),
  list(res_softrsr,   "Soft RSR penalty")
)

beta_rsr_df <- do.call(rbind, lapply(rows_rsr, function(x) {
  r <- x[[1]]; nm <- x[[2]]
  if (is.null(r) || is.null(r$beta_hat)) return(NULL)
  data.frame(Model     = nm,
             Intercept = round(r$beta_hat[1], 3),
             Water     = round(r$beta_hat[2], 3),
             rho       = round(r$rho_opt %||% 1, 3),
             phi       = round(r$phi, 3),
             stringsAsFactors = FALSE)
}))

if (!is.null(beta_rsr_df) && nrow(beta_rsr_df) > 0) {
  kable(beta_rsr_df, format = "latex", booktabs = TRUE, escape = FALSE,
        caption = paste0("Estimated covariate coefficients under RSR-constrained models. ",
                         "The water coefficient is now negative and physically consistent: ",
                         "the model assigns lower SIF to water pixels, ",
                         "consistent with near-zero photosynthetic activity over water."),
        label = "betas-rsr")
} else {
  cat("RSR SIF results not available.")
}
```

Under RSR, the water coefficient is negative and of plausible magnitude,
consistent with the prior expectation from the literature and with lower
raw SIF values observed over water soundings. The intercept estimates mean
SIF for vegetated pixels, which is also consistent with typical SIF values
for the Boston region in summer.

## Why not simply exclude water pixels?

One alternative is to set $\beta_\text{water} = 0$ and exclude water pixels
from contributing to covariate adjustment. The difficulty is that this
requires the spatial prior to account for all variation at land-water
boundaries without any covariate information. With limited sounding coverage
over water, the prior dominates in those pixels and boundary artefacts can
appear. RSR uses the covariate to explain the land-water mean difference and
reserves the spatial prior for residual variation, which makes better use of
the available information.

## Soft RSR penalty

Hard RSR enforces the orthogonality constraint exactly. A soft alternative
penalises departures from orthogonality with a scale parameter $\lambda$:

$$
\text{penalty}(\lambda) = \frac{\lambda}{2} \| X^{(o)\top} A r^{(\ell)} \|^2.
$$

When $\lambda \to \infty$ this recovers hard RSR; when $\lambda = 0$ the
constraint is absent. Cross-validation over a grid of $\lambda$ values
selects a moderately penalised solution.

```{r soft-rsr-plot, fig.cap="Cross-validation MSE surface for the soft RSR penalty parameter $\\alpha$ at four values of $\\rho$. Each curve shows the CV MSE as a function of $\\log\\alpha$; the selected value (star) prefers moderate penalisation between the unconstrained ($\\alpha\\to 0$) and hard-RSR ($\\alpha\\to\\infty$) extremes.", fig.height=3}
if (!is.null(res_softrsr) && !is.null(res_softrsr$grid)) {
  g <- res_softrsr$grid
  g$rho_label <- factor(sprintf("rho=%.2f", g$rho))
  best <- g[which.min(g$cv_mse), ]

  ggplot(g, aes(x = log_alpha, y = cv_mse, colour = rho_label)) +
    geom_line(linewidth = 0.8) +
    geom_point(data = best, shape = 8, size = 3, colour = "black") +
    scale_colour_brewer(palette = "RdYlBu", name = expression(rho)) +
    labs(x     = expression(log~alpha),
         y     = "CV MSE",
         title = "Soft RSR: CV surface over penalty scale") +
    theme_minimal(base_size = 10)
} else {
  plot.new()
  text(0.5, 0.5, "Soft RSR results not available.\nRun soft RSR script first.", cex = 1)
}
```

\newpage

# Landcover-Aware Neighbourhood Weights {#sec:neighbors}

## Motivation

The standard SAR prior uses queen adjacency with equal weights, treating
all edges between neighbouring pixels symmetrically. An alternative is to
reduce weights on edges crossing land-water boundaries, on the grounds that
SIF is expected to vary more sharply at such interfaces. We evaluated four
neighbourhood variants alongside the standard queen adjacency baseline:
rook adjacency (no diagonal neighbours), land-cover downweighted queen
adjacency, and two parameterisations of cross-boundary weight reduction.

## Results

```{r neighbor-table}
res_base    <- load_result("results_water_rsr_rho1")
res_rook    <- load_result("results_neighbor_rook")
res_lc      <- load_result("results_neighbor_lc")
res_lc_eps  <- load_result("results_neighbor_lc_eps")
res_lc_alph <- load_result("results_neighbor_lc_alpha")

rows_nb <- list(
  list(res_base,    "Queen, equal weights (baseline)"),
  list(res_rook,    "Rook, equal weights"),
  list(res_lc,      "Queen + LC downweight"),
  list(res_lc_eps,  "Queen + LC $\\epsilon$-weight"),
  list(res_lc_alph, "Queen + LC $\\alpha$-weight")
)

nb_df <- do.call(rbind, lapply(rows_nb, function(x) {
  r <- x[[1]]; nm <- x[[2]]
  if (is.null(r)) return(NULL)
  data.frame(Model    = nm,
             phi      = round(r$phi, 3),
             tau      = round(1/r$phi, 2),
             RMSE     = round(r$rmse %||% NA_real_, 5),
             R2       = round(r$r2   %||% NA_real_, 4),
             Coverage = round(r$coverage_95_obs %||% NA_real_, 4),
             stringsAsFactors = FALSE)
}))

if (!is.null(nb_df) && nrow(nb_df) > 0) {
  kable(nb_df, format = "latex", booktabs = TRUE, escape = FALSE,
        caption = paste0("Predictive performance under alternative neighbourhood weight schemes ",
                         "on the semi-synthetic albedo data. Differences in RMSE and R$^2$ ",
                         "are small across all variants, suggesting that once water proportion ",
                         "is included as a covariate, the specific neighbourhood structure ",
                         "has limited additional impact."),
        label = "neighbor-table")
} else {
  cat("Neighbour weight results not available.")
}
```

```{r neighbor-maps, fig.cap="Posterior mean albedo under equal-weight queen adjacency (left) and landcover-downweighted adjacency (right). Spatial patterns are nearly indistinguishable, confirming that the water covariate absorbs most of the boundary information.", fig.height=3.5}
if (!is.null(res_base) && !is.null(res_lc) &&
    !is.null(res_base$posterior_mean) && !is.null(res_lc$posterior_mean)) {
  target_sf$val_base <- expand_to_grid(res_base$posterior_mean, target_idx, n_grid)
  target_sf$val_lc   <- expand_to_grid(res_lc$posterior_mean,   target_idx, n_grid)
  lims <- range(c(target_sf$val_base, target_sf$val_lc), na.rm = TRUE)

  p1 <- ggplot(target_sf) +
    geom_sf(aes(fill = val_base), colour = NA) +
    scale_fill_viridis_c(limits = lims, na.value = "grey90", name = "Albedo") +
    labs(title = "Queen (equal weights)") +
    theme_minimal(base_size = 9) +
    theme(axis.text = element_blank(), axis.ticks = element_blank(),
          axis.title = element_blank(), panel.grid = element_blank())

  p2 <- ggplot(target_sf) +
    geom_sf(aes(fill = val_lc), colour = NA) +
    scale_fill_viridis_c(limits = lims, na.value = "grey90", name = "Albedo") +
    labs(title = "Queen + LC downweight") +
    theme_minimal(base_size = 9) +
    theme(axis.text = element_blank(), axis.ticks = element_blank(),
          axis.title = element_blank(), panel.grid = element_blank())

  print(p1 | p2)
} else {
  plot.new()
  text(0.5, 0.5, "Neighbour posterior results not available.", cex = 1)
}
```

The landcover-aware weights produce negligible improvement in predictive
performance on the albedo data. This is consistent with the expectation
that, once water proportion is included as a covariate, the model already
accounts for the mean difference between land and water pixels; the
neighbourhood structure then affects only residual spatial smoothing, which
is relatively insensitive to land-water boundaries once the covariate is
present. On this basis, we use equal-weight queen adjacency as the primary
neighbourhood specification.

\newpage

# Forward Operator Sensitivity {#sec:gA}

## Motivation

The primary analysis uses a uniform area-intersection forward operator,
treating all parts of a sounding footprint as equally informative. The
OCO-2 sensor response may in practice taper toward footprint edges,
motivating a Gaussian-weighted alternative:
$$
A_{ij} \;\propto\; \int_{D^{(o)}_i \cap D^{(\ell)}_j} g_i(s)\,\mathrm{d}s,
$$
where $g_i$ is an affine-mapped isotropic Gaussian with scale $\tau$ in
canonical $[-1,1]^2$ footprint space. The value $\tau = 1/3$ places the
footprint edge at approximately one standard deviation.

Since the true OCO-2 sensor response function for target-mode retrievals is
not precisely known, and since $\tau$ cannot be tuned without ground truth,
we treat this as a sensitivity analysis rather than an alternative primary
model.

## 2x2 contingency experiment

We generated synthetic albedo observations under each forward operator
and fitted under each forward operator, yielding a $2\times2$ contingency
table. Diagonal cells are correctly specified; off-diagonal cells are
misspecified.

```{r contingency-table}
r_uu <- load_result("results_contingency_uu")
r_ug <- load_result("results_contingency_ug")
r_gu <- load_result("results_contingency_gu")
r_gg <- load_result("results_contingency_gg")

if (!any(sapply(list(r_uu, r_ug, r_gu, r_gg), is.null))) {
  cont_df <- data.frame(
    `Generating $A$` = c("Uniform", "Uniform", "Gaussian", "Gaussian"),
    `Fitting $A$`    = c("Uniform", "Gaussian", "Uniform", "Gaussian"),
    RMSE = round(c(r_uu$rmse, r_ug$rmse, r_gu$rmse, r_gg$rmse), 5),
    `$R^2$` = round(c(r_uu$r2, r_ug$r2, r_gu$r2, r_gg$r2), 4),
    phi  = round(c(r_uu$tuned$phi, r_ug$tuned$phi,
                   r_gu$tuned$phi, r_gg$tuned$phi), 3),
    check.names = FALSE
  )
  kable(cont_df, format = "latex", booktabs = TRUE, escape = FALSE,
        caption = paste0("$2\\times 2$ contingency table comparing generating vs fitting ",
                         "forward operator on semi-synthetic albedo data. ",
                         "Diagonal entries (correctly specified) outperform off-diagonal ",
                         "entries (misspecified). Misspecification is asymmetric: ",
                         "fitting uniform $A$ to Gaussian-generated data (RMSE ",
                         round(r_gu$rmse, 5), ") is worse than the reverse (",
                         round(r_ug$rmse, 5), "), ",
                         "because the Gaussian operator concentrates mass near footprint ",
                         "centres, making uniform averaging a poorer approximation."),
        label = "contingency")
} else {
  cat("Contingency results not yet available. Run run_g_A_contingency.R first.")
}
```

The correctly specified diagonal entries outperform both off-diagonal
entries. Because $\tau$ cannot be estimated without ground truth, and
because the uniform operator is validated under the semi-synthetic
experiment, we retain the uniform forward operator as the primary model.

\newpage

# Hyperparameter Tuning: Cross-Validation, Spatial Blocking, and Maximum Likelihood {#sec:tuning}

## Overview

The smoothing parameter $\tau = 1/\phi$ is selected by 10-fold
cross-validation (CV) minimising held-out prediction error. This section
examines two alternative tuning approaches — spatially blocked CV and
marginal maximum likelihood (ML) — and explains why random CV is preferred.

## Spatial blocking

```{r block-map, fig.cap="Spatial fold assignments for the blocked cross-validation experiment. Ten geographically contiguous blocks were formed by $k$-means clustering of sounding centroids.", fig.height=4}
blocked_folds <- d_sif$blocked_folds %||% goebel2026::setup_albedo$blocked_folds
soundings_sf  <- d_shared$soundings_proj

if (!is.null(blocked_folds) && !is.null(soundings_sf)) {
  soundings_sf$block <- factor(blocked_folds)
  ggplot(soundings_sf) +
    geom_sf(aes(fill = block), colour = NA, alpha = 0.85) +
    scale_fill_brewer(palette = "Paired", name = "Block") +
    labs(title = "Spatially blocked CV folds ($k$-means, $k=10$)") +
    theme_minimal(base_size = 9) +
    theme(axis.text = element_blank(), axis.ticks = element_blank(),
          panel.grid = element_blank())
} else {
  plot.new(); text(0.5, 0.5, "blocked_folds not available", cex = 1)
}
```

```{r blocking-table}
res_cv_alb  <- load_result("results_water_rho1")
res_blk_alb <- load_result("results_cv_blocked")
res_cv_sif  <- load_result("results_sif_canonical")
res_blk_sif <- load_result("results_sif_cv_blocked")

rows_blk <- list(
  list(res_cv_alb,  "Albedo", "Random CV"),
  list(res_blk_alb, "Albedo", "Blocked CV"),
  list(res_cv_sif,  "SIF",    "Random CV"),
  list(res_blk_sif, "SIF",    "Blocked CV")
)

blk_df <- do.call(rbind, lapply(rows_blk, function(x) {
  r <- x[[1]]; ds <- x[[2]]; mt <- x[[3]]
  if (is.null(r)) return(NULL)
  data.frame(Dataset  = ds,
             Method   = mt,
             phi      = round(r$phi, 4),
             tau      = round(1/r$phi, 2),
             RMSE     = round(r$rmse %||% NA_real_, 5),
             `$R^2$`  = round(r$r2   %||% NA_real_, 4),
             check.names = FALSE,
             stringsAsFactors = FALSE)
}))

if (!is.null(blk_df) && nrow(blk_df) > 0) {
  kable(blk_df, format = "latex", booktabs = TRUE, escape = FALSE,
        caption = paste0("Smoothing parameter selected by random vs spatially blocked CV. ",
                         "Blocked CV selects substantially smaller $\\phi$ (stronger smoothing) ",
                         "on both datasets, with the discrepancy approximately $14\\times$ ",
                         "for albedo and $55\\times$ for SIF."),
        label = "blocking-table")
} else {
  cat("Blocking results not available.")
}
```

Spatially blocked CV selects much smaller $\phi$ than random CV on both
datasets. This reflects the OCO-2 target-mode observation geometry: sounding
footprints overlap densely in the central domain, so any geographic
partition creates train-test splits with radically different overlap
densities. In a held-out block, the spatial prior cannot extrapolate across
the gap from training soundings, so the CV objective provides no gradient
favouring smoothing and $\phi$ collapses toward zero.

The standard justification for spatial blocking — preventing autocorrelation
leakage between training and test folds — does not apply to this model.
There is no spatial noise term: spatial structure enters only through the
SAR prior, which is integrated out. Leakage between folds is therefore not
a concern, and random CV is the appropriate evaluation criterion.

## ML vs CV: likelihood surface analysis

Figure \ref{fig:likelihood-surface} shows ML likelihood and CV MSE profiles
as a function of $\log\phi$ at four fixed values of $\rho$. Both curves are
computed on the no-covariate, no-RSR model to ensure direct comparability
between objectives.

```{r likelihood-surface, fig.cap="ML likelihood profiles (red) and CV MSE profile at $\\rho=0.99$ (blue, bottom-right panel only), both scaled to $[0,1]$ for comparability. Stars mark ML optima; dashed verticals mark CV-selected $\\phi$ at each $\\rho$. At low $\\rho$, ML prefers more smoothing than CV. At $\\rho=0.95$ the two methods agree closely. At $\\rho=0.99$, ML collapses to small $\\phi$ due to distortion of the likelihood surface near the intrinsic prior boundary.", fig.height=6}
ml_grid_file <- "likelihood_surface_ml_grid.rds"
cv_grid_file <- "likelihood_surface_cv_grid.rds"
cv_opts_file <- "likelihood_surface_cv_opts.rds"

if (file.exists(ml_grid_file) && file.exists(cv_opts_file)) {
  ml_grid <- readRDS(ml_grid_file)
  cv_grid <- if (file.exists(cv_grid_file)) readRDS(cv_grid_file) else NULL
  cv_opts <- readRDS(cv_opts_file)

  rho_values <- sort(unique(ml_grid$rho))
  rho_cv     <- if (!is.null(cv_grid)) cv_grid$rho[1] else 0.99

  ml_grid <- do.call(rbind, lapply(split(ml_grid, ml_grid$rho), function(d) {
    d$ll_norm   <- d$ll - max(d$ll, na.rm = TRUE)
    d$ll_scaled <- (d$ll_norm - min(d$ll_norm, na.rm = TRUE)) /
                   diff(range(d$ll_norm, na.rm = TRUE))
    d
  }))

  ml_opt <- do.call(rbind, lapply(rho_values, function(rho) {
    sub <- ml_grid[ml_grid$rho == rho & is.finite(ml_grid$ll), ]
    sub[which.max(sub$ll), c("rho", "phi", "log_phi", "ll")]
  }))

  panels <- lapply(rho_values, function(rho) {
    ml_sub <- ml_grid[ml_grid$rho == rho & is.finite(ml_grid$ll_scaled), ]
    ml_o   <- ml_opt[ml_opt$rho == rho, ]
    cv_o   <- cv_opts[cv_opts$rho == rho, ]
    has_cv <- (!is.null(cv_grid) && rho == rho_cv)

    p <- ggplot() +
      geom_line(data = ml_sub, aes(x = log_phi, y = ll_scaled),
                colour = "tomato", linewidth = 0.8) +
      geom_point(data = ml_o, aes(x = log_phi, y = 1),
                 colour = "tomato", shape = 8, size = 2.5)

    if (has_cv) {
      cv_sub <- cv_grid[is.finite(cv_grid$cv_mse), ]
      cv_sub$cv_scaled <- 1 - (cv_sub$cv_mse - min(cv_sub$cv_mse, na.rm=TRUE)) /
                              diff(range(cv_sub$cv_mse, na.rm=TRUE))
      p <- p + geom_line(data = cv_sub, aes(x = log_phi, y = cv_scaled),
                         colour = "steelblue", linewidth = 0.8)
    }

    p +
      geom_vline(xintercept = ml_o$log_phi,     colour = "tomato",
                 linetype = "dashed", linewidth = 0.5) +
      geom_vline(xintercept = log(cv_o$phi),     colour = "steelblue",
                 linetype = "dashed", linewidth = 0.5) +
      annotate("text", x = ml_o$log_phi,    y = 0.05,
               label = sprintf("ML\n%.3f", ml_o$phi),
               hjust = 1.1, size = 2.4, colour = "tomato") +
      annotate("text", x = log(cv_o$phi), y = 0.18,
               label = sprintf("CV\n%.3f", cv_o$phi),
               hjust = -0.1, size = 2.4, colour = "steelblue") +
      scale_y_continuous(limits = c(0, 1), labels = NULL) +
      labs(x = expression(log~phi), y = NULL,
           title = bquote(rho == .(rho)),
           subtitle = if (has_cv) "Red = ML  |  Blue = CV (scaled)"
                      else        "Red = ML  |  Blue dashed = CV opt") +
      theme_minimal(base_size = 9) +
      theme(panel.grid.minor = element_blank())
  })

  print(
    (panels[[1]] | panels[[2]]) / (panels[[3]] | panels[[4]]) +
      plot_annotation(caption = "SIF data, no covariates, no RSR, no $R^{-1}$.")
  )
} else {
  cat("Likelihood surface cache files not found.",
      "Run plot\\_likelihood\\_surface.R to generate them.")
}
```

The ML optimum shifts systematically with $\rho$. At moderate $\rho$,
ML prefers more smoothing than CV; at $\rho \approx 0.95$ the two methods
agree closely; at $\rho = 0.99$ ML collapses to very small $\phi$. This
collapse arises because the log-determinant term $\tfrac{1}{2}\log|Q(\rho)|$
in the ML objective diverges as $Q$ becomes increasingly ill-conditioned
near the intrinsic prior boundary, distorting the likelihood surface. The CV
objective is unaffected by prior geometry and maintains a stable interior
minimum across all $\rho$ values.

```{r ml-cv-table}
results_for_table <- list(
  list("Random CV, RSR + $R^{-1}$ (canonical)", "SIF",
       load_result("results_sif_canonical")),
  list("Blocked CV, RSR + $R^{-1}$", "SIF",
       load_result("results_sif_cv_blocked")),
  list("ML, $\\rho$ free", "SIF",
       load_result("results_sif_water_ml")),
  list("ML, $\\rho=0.95$ fixed", "SIF",
       load_result("results_sif_rho095_ml")),
  list("CV, $\\rho=0.95$ fixed", "SIF",
       load_result("results_sif_rho095_cv")),
  list("Random CV, RSR (canonical)", "Albedo",
       load_result("results_water_rsr_rho1")),
  list("Blocked CV, RSR", "Albedo",
       load_result("results_cv_blocked")),
  list("ML, $\\rho$ free", "Albedo",
       load_result("results_water_ml"))
)

tbl_df <- do.call(rbind, lapply(results_for_table, function(x) {
  nm <- x[[1]]; ds <- x[[2]]; r <- x[[3]]
  if (is.null(r)) return(NULL)
  data.frame(Method  = nm,
             Dataset = ds,
             rho     = round(r$rho_opt %||% 1, 3),
             phi     = round(r$phi, 4),
             tau     = round(1/r$phi, 2),
             stringsAsFactors = FALSE)
}))

if (!is.null(tbl_df) && nrow(tbl_df) > 0) {
  kable(tbl_df, format = "latex", booktabs = TRUE, escape = FALSE,
        caption = paste0("Summary of $\\phi$ and $\\tau = 1/\\phi$ selected by each tuning ",
                         "method on SIF and albedo data. Random CV and ML agree well at ",
                         "$\\rho = 0.95$; blocked CV and free-$\\rho$ ML diverge due to ",
                         "the geometric and likelihood-surface issues described above."),
        label = "ml-cv-table")
} else {
  cat("Tuning comparison results not available.")
}
```
