---
title: "Supplementary Materials: Bayesian Downscaling of Satellite-Derived Solar-Induced Chlorophyll Fluorescence"
author: "Russell Goebel, Leeza Moldavchuk, Taylor S. Jones, Jonathan Dooley, Lucy R. Hutyra, Luis Carvalho"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    latex_engine: pdflatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  out.width = "\\textwidth"
)

library(goebel2026)
library(ggplot2)
library(patchwork)
library(dplyr)
library(knitr)
library(scales)
library(sf)
library(gstat)

# Shared formatting helpers
fmt2 <- function(x) sprintf("%.2f", x)
fmt3 <- function(x) sprintf("%.3f", x)
fmt4 <- function(x) sprintf("%.4f", x)
fmt1 <- function(x) sprintf("%.1f", x)
pct  <- function(x) sprintf("%.1f%%", 100 * x)

# Load all result objects fresh from package (avoids stale session cache)
.pkg_get <- function(nm) get(nm, envir = asNamespace("goebel2026"))

r <- list(
  # Albedo ablation
  ols             = .pkg_get("results_ols_baseline"),
  no_cov          = .pkg_get("results_no_cov_rho1"),
  water           = .pkg_get("results_water_rho1"),
  water_rsr       = .pkg_get("results_water_rsr_rho1"),
  water_rsr_cv    = .pkg_get("results_water_rsr_rho_cv"),
  water_rho_cv    = .pkg_get("results_water_rho_cv"),
  # SIF
  sif             = .pkg_get("results_sif_canonical"),
  sif_naive       = .pkg_get("results_sif_naive"),
  sif_rsr_rho1    = .pkg_get("results_sif_rsr_rho1"),
  # Tuning
  cv_blocked      = .pkg_get("results_cv_blocked"),
  water_ml        = .pkg_get("results_water_ml"),
  sif_cv_blocked  = .pkg_get("results_sif_cv_blocked"),
  sif_water_ml    = .pkg_get("results_sif_water_ml"),
  sif_rho095_ml   = .pkg_get("results_sif_rho095_ml"),
  sif_rho095_cv   = .pkg_get("results_sif_rho095_cv"),
  # Neighbours
  nb_rook         = .pkg_get("results_neighbor_rook"),
  nb_lc_eps       = .pkg_get("results_neighbor_lc_eps"),
  nb_lc_alpha     = .pkg_get("results_neighbor_lc_alpha"),
  sif_nb_rook     = .pkg_get("results_sif_neighbor_rook"),
  sif_nb_lc_eps   = .pkg_get("results_sif_neighbor_lc_eps"),
  sif_nb_lc_alpha = .pkg_get("results_sif_neighbor_lc_alpha"),
  # Forward operator
  alb_gA_tau05    = .pkg_get("results_albedo_gA_tau05"),
  alb_gA_tau033   = .pkg_get("results_albedo_gA_tau033"),
  alb_gA_tau02    = .pkg_get("results_albedo_gA_tau02"),
  alb_gA_tau01    = .pkg_get("results_albedo_gA_tau01"),
  alb_gA_cent     = .pkg_get("results_albedo_gA_centroid"),
  sif_gA_tau05    = .pkg_get("results_sif_gA_tau05"),
  sif_gA_tau033   = .pkg_get("results_sif_gA_tau033"),
  sif_gA_tau02    = .pkg_get("results_sif_gA_tau02"),
  sif_gA_tau01    = .pkg_get("results_sif_gA_tau01"),
  sif_gA_tau001   = .pkg_get("results_sif_gA_tau001"),
  # Kriging
  krig_alb        = .pkg_get("results_kriging_albedo"),
  krig_sif        = .pkg_get("results_kriging_sif"),
  # Soft RSR
  softrsr         = .pkg_get("results_softrsr_albedo"),
  # Contingency
  ct_uu           = .pkg_get("results_contingency_uu"),
  ct_ug           = .pkg_get("results_contingency_ug"),
  ct_gu           = .pkg_get("results_contingency_gu"),
  ct_gg           = .pkg_get("results_contingency_gg")
)

# Likelihood surfaces
ls_alb <- .pkg_get("likelihood_surface_albedo")
ls_sif <- .pkg_get("likelihood_surface_sif")

# Safe rho extraction -- fixed-rho results don't store rho_opt
get_rho <- function(x, default = 1.0) {
  v <- x$rho_opt
  if (is.null(v) || length(v) == 0) default else v
}

# Safe field extraction -- returns NA if field is NULL/missing
gf <- function(x, field, idx = NULL) {
  v <- x[[field]]
  if (is.null(v) || length(v) == 0) return(NA_real_)
  if (!is.null(idx)) v <- v[idx]
  if (length(v) == 0) NA_real_ else v
}

# Safe field extraction (multi-name fallback)
get_field <- function(x, ...) {
  fields <- c(...)
  for (f in fields) {
    v <- x[[f]]
    if (!is.null(v) && length(v) > 0) return(v)
  }
  NA_real_
}
```

\newpage

# Hyperparameter Tuning {#sec:tuning}

The smoothing parameter $\phi$ controls the signal-to-noise ratio of the
spatial random effect and must be selected from data. We compare three
approaches: (1) random $K$-fold cross-validation (random CV), which assigns
soundings to folds uniformly at random; (2) spatially blocked cross-validation
(blocked CV), which assigns soundings to contiguous spatial blocks; and (3)
marginal likelihood maximisation (ML). Throughout this section we fix
$\rho = 0.95$ and include no covariates or RSR constraint, so that differences
in selected $\phi$ reflect the tuning method alone.

**Spatial blocking criterion.** For blocked CV, soundings are assigned to
folds by $k$-means clustering on their centroid coordinates, with $k = 10$.
This produces spatially contiguous blocks whose sizes reflect the local
sounding density — denser regions (the centre of the target-mode acquisition)
yield larger blocks. Figure 1 shows the resulting fold
assignments overlaid on the sounding geometries, coloured by fold index.
The concentration of soundings in the central strip means that a single
held-out block may remove a disproportionately large and informationally dense
region from the training set, which we conjecture contributes to the
divergent $\phi$ selection documented below.

```{r fold_map, fig.cap="Spatial fold assignments for blocked (left) and random (right) 10-fold cross-validation. Each sounding is coloured by its fold index. Blocked folds form contiguous spatial regions; random folds are scattered uniformly. The central high-density strip falls predominantly within one or two blocked folds, meaning that holding out a single block may remove a disproportionately informative set of soundings from the training set.", fig.height=4, fig.width=9}

soundings_sf  <- goebel2026::setup_shared$soundings_proj
folds_blocked <- goebel2026::setup_albedo$blocked_folds
set.seed(2026L)
folds_random  <- sample(rep_len(1:10, nrow(soundings_sf)))

soundings_sf$fold_blocked <- factor(folds_blocked)
soundings_sf$fold_random  <- factor(folds_random)

bb <- sf::st_bbox(soundings_sf)

make_fold_plot <- function(sf_obj, fold_col, title) {
  ggplot(sf_obj) +
    geom_sf(aes(fill = .data[[fold_col]]), color = NA, alpha = 0.8) +
    scale_fill_manual(
      values = setNames(scales::hue_pal()(10), as.character(1:10)),
      name = "Fold", guide = guide_legend(ncol = 2)
    ) +
    coord_sf(xlim = c(bb["xmin"], bb["xmax"]),
             ylim = c(bb["ymin"], bb["ymax"]),
             expand = FALSE) +
    labs(title = title) +
    theme_void(base_size = 9) +
    theme(legend.position  = "right",
          plot.title       = element_text(size = 10, hjust = 0.5))
}

p_blk <- make_fold_plot(soundings_sf, "fold_blocked", "Blocked CV")
p_rnd <- make_fold_plot(soundings_sf, "fold_random",  "Random CV")
p_blk | p_rnd
```

## Likelihood and CV objective profiles

Figure 2 shows the ML log-likelihood and CV MSE
profiles as functions of $\log\phi$ for both albedo (left) and SIF (right).
All three objectives are scaled to $[0, 1]$ and oriented so that higher is
better.

For both responses, the ML and random CV profiles peak at similar values of
$\phi$: the ratio of ML to random CV optima is
`r fmt2(ls_alb$ml_phi_opt / ls_alb$cv_phi_random)`x for albedo and
`r fmt2(ls_sif$ml_phi_opt / ls_sif$cv_phi_random)`x for SIF, indicating
strong agreement between the two methods. Both profiles have broad, flat
peaks, suggesting that predictions are relatively insensitive to the exact
value of $\phi$ in this region.

Spatially blocked CV selects a substantially different $\phi$: its optimum is
`r fmt2(ls_alb$cv_phi_blocked / ls_alb$cv_phi_random)`x larger than the
random CV optimum for albedo and
`r fmt2(ls_sif$cv_phi_blocked / ls_sif$cv_phi_random)`x larger for SIF.
One plausible explanation for this discrepancy is that removing a contiguous
spatial block disproportionately removes the most densely observed region of
the domain, creating a train/test density mismatch that may not reflect
generalisation error at the actual observation density. Under this
interpretation, the held-out block is farther from its training neighbours
than a typical held-out sounding would be, possibly inducing heavier smoothing
than the data warrant. Random CV, by contrast, holds out scattered soundings
whose spatial context in the training set more closely mirrors the full
dataset, which may explain why its selected $\phi$ agrees more closely with
the marginal likelihood optimum.

```{r likelihood_surface, fig.cap="ML log-likelihood (red) and CV MSE profiles (random CV: blue; blocked CV: green) as functions of $\\log\\phi$, for albedo (left) and SIF (right). The ML curve is scaled to $[0,1]$ globally; CV curves are normalized within a $\\pm 6$ log-unit window around each optimum so that peak shape is visible. All curves are oriented so that higher is better. Dashed vertical lines indicate the optimum for each method. Random CV and ML optima are in close agreement; blocked CV selects a substantially larger $\\phi$ in both cases.", fig.height=4, fig.width=9}

make_surface_panel <- function(surf, title, ylim = c(-0.05, 0.05)) {
  lp  <- surf$log_phi_seq

  ml  <- surf$ml_ll
  ml  <- (ml  - max(ml,  na.rm = TRUE)) /
          abs(diff(range(ml,  na.rm = TRUE)))

  # Normalize CV curves relative to variation within ±6 log-phi units of the
  # optimum so the peak shape is visible rather than compressed by extreme tails.
  local_norm <- function(x, lp) {
    opt_lp  <- lp[which.min(x)]
    window  <- abs(lp - opt_lp) <= 6
    x_min   <- min(x, na.rm = TRUE)
    x_range <- abs(diff(range(x[window], na.rm = TRUE)))
    if (x_range == 0) x_range <- 1
    -(x - x_min) / x_range
  }

  cvr <- local_norm(surf$cv_mse_random,  lp)
  cvb <- local_norm(surf$cv_mse_blocked, lp)

  df <- data.frame(
    log_phi = rep(lp, 3),
    value   = c(ml, cvr, cvb),
    method  = rep(c("ML", "CV random", "CV blocked"), each = length(lp))
  )
  df$method <- factor(df$method, levels = c("ML", "CV random", "CV blocked"))

  opts <- data.frame(
    method  = factor(c("ML", "CV random", "CV blocked"),
                     levels = c("ML", "CV random", "CV blocked")),
    log_phi = log(c(surf$ml_phi_opt, surf$cv_phi_random, surf$cv_phi_blocked))
  )

  subtitle_text <- sprintf(
    "phi_opt -- ML: %.3g  |  Rand: %.3g  |  Block: %.3g",
    surf$ml_phi_opt, surf$cv_phi_random, surf$cv_phi_blocked
  )

  ggplot(df[is.finite(df$value), ],
         aes(x = log_phi, y = value, color = method)) +
    geom_line(linewidth = 0.9) +
    geom_vline(data = opts,
               aes(xintercept = log_phi, color = method),
               linetype = "dashed", linewidth = 0.6) +
    scale_color_manual(
      values = c("ML"         = "tomato",
                 "CV random"  = "steelblue",
                 "CV blocked" = "darkgreen"),
      name = NULL) +
    scale_y_continuous(limits = c(NA, 0.05)) +
    coord_cartesian(ylim = ylim, xlim = c(-6, 13)) +
    labs(x        = expression(log~phi),
         y        = "Scaled objective (higher = better)",
         title    = title,
         subtitle = subtitle_text) +
    theme_minimal(base_size = 10) +
    theme(legend.position   = "bottom",
          panel.grid.minor  = element_blank(),
          plot.subtitle     = element_text(size = 8, color = "grey40"))
}

p_alb <- make_surface_panel(ls_alb, "Albedo", ylim = c(-1.05, 0.05))
p_sif <- make_surface_panel(ls_sif, "SIF",    ylim = c(-1.05, 0.05))
p_alb | p_sif
```

## Comparison with covariate model

The results above use no covariates or RSR constraint. When the water
proportion covariate and RSR are included (as in the main analysis),
the tuning methods still disagree. Table 1 shows CV
results with RSR: blocked CV selects $\phi$ approximately
`r fmt2(gf(r$water_rsr,"phi") / gf(r$cv_blocked,"phi"))`x smaller than
random CV for albedo and
`r fmt2(gf(r$sif_rsr_rho1,"phi") / gf(r$sif_cv_blocked,"phi"))`x smaller
for SIF, with coverage collapsing from
`r pct(gf(r$water_rsr,"coverage_95_obs"))` to
`r pct(gf(r$cv_blocked,"coverage_95_obs"))` for albedo.

Table 2 shows all three methods without RSR or
covariates, at fixed $\rho=0.95$ — the same conditions as
Figure 1. Random CV and ML agree closely (ratio
`r fmt2(ls_alb$ml_phi_opt / ls_alb$cv_phi_random)`x for albedo,
`r fmt2(ls_sif$ml_phi_opt / ls_sif$cv_phi_random)`x for SIF), while
blocked CV selects a substantially larger $\phi$ in both cases. ML cannot
be combined with RSR since RSR is a posterior constraint with no natural
analogue in the marginal likelihood objective; the
`r fmt2(gf(r$sif_rho095_cv,"phi") / gf(r$sif_rho095_ml,"phi"))`x ratio
between ML and random CV at fixed $\rho=0.95$ with RSR+covariates
(not shown) is consistent with the no-RSR comparison.

Crucially, $\hat\beta_\text{water}$ is stable across both tables and both
responses (range:
`r fmt4(min(sapply(list(r$water_rsr, r$cv_blocked, r$water_ml), gf, "beta_hat", idx=2), na.rm=TRUE))`
to
`r fmt4(max(sapply(list(r$water_rsr, r$cv_blocked, r$water_ml), gf, "beta_hat", idx=2), na.rm=TRUE))`
for albedo;
`r fmt4(min(sapply(list(r$sif_rsr_rho1, r$sif_cv_blocked, r$sif_rho095_ml), gf, "beta_hat", idx=2), na.rm=TRUE))`
to
`r fmt4(max(sapply(list(r$sif_rsr_rho1, r$sif_cv_blocked, r$sif_rho095_ml), gf, "beta_hat", idx=2), na.rm=TRUE))`
for SIF), confirming that the tuning method does not affect the
substantive inference. Because random CV directly optimises prediction
accuracy, produces near-nominal coverage in the semi-synthetic
validation, and is consistent with ML at comparable model specifications,
we prefer it for the main analysis.

```{r tuning_cv}
tuning_cv_alb <- list(r$water_rsr, r$cv_blocked)
tuning_cv_sif <- list(r$sif_rsr_rho1, r$sif_cv_blocked)

tab_cv <- data.frame(
  Response = c("Albedo", "Albedo", "SIF", "SIF"),
  Method   = rep(c("Random CV", "Blocked CV"), 2),
  rho      = c(sapply(tuning_cv_alb, get_rho),
               sapply(tuning_cv_sif, get_rho)),
  phi      = c(sapply(tuning_cv_alb, gf, "phi"),
               sapply(tuning_cv_sif, gf, "phi")),
  RMSE     = c(sapply(tuning_cv_alb, gf, "rmse"),
               NA_real_, NA_real_),
  Coverage = c(sapply(tuning_cv_alb, gf, "coverage_95_obs"),
               NA_real_, NA_real_),
  mean_SE  = c(NA_real_, NA_real_,
               sapply(tuning_cv_sif, function(x)
                 mean(x$posterior_se, na.rm = TRUE)))
)

kable(tab_cv,
  format    = "pipe",
  digits    = 4,
  caption   = "CV tuning methods with RSR constraint. Albedo results include RMSE and coverage (ground truth known); SIF reports mean posterior SE.",
  col.names = c("Response", "Method", "rho", "phi",
                "RMSE", "Coverage (95%)", "Mean SE"))
```

```{r tuning_nrsr}
tab_nrsr <- data.frame(
  Response = c("Albedo", "Albedo", "Albedo", "SIF", "SIF", "SIF"),
  Method   = rep(c("Random CV", "Blocked CV", "ML"), 2),
  rho      = 0.95,
  phi      = c(ls_alb$cv_phi_random,
               ls_alb$cv_phi_blocked,
               ls_alb$ml_phi_opt,
               ls_sif$cv_phi_random,
               ls_sif$cv_phi_blocked,
               ls_sif$ml_phi_opt)
)

kable(tab_nrsr,
  format    = "pipe",
  digits    = 4,
  caption   = "All three tuning methods without RSR constraint or covariates, at fixed rho=0.95. These are the phi optima shown in Figure 2. Random CV and ML agree closely for both responses; blocked CV selects a substantially larger phi due to the dense OCO-2 overlap structure.",
  col.names = c("Response", "Method", "rho", "phi (opt)"))
```

\newpage

# Neighbour Weight Sensitivity {#sec:neighbours}

The SAR prior is defined through a neighbourhood matrix $W$. We consider
four specifications, all using queen adjacency as the base structure.
Let $\mathcal{N}(i)$ denote the queen-adjacency neighbours of cell $i$,
$n_i = |\mathcal{N}(i)|$, $w_i \in [0,1]$ the water proportion of cell $i$,
and $\ell_{ij} = \mathbf{1}\{w_i > 0.5 \neq w_j > 0.5\}$ a binary indicator
that the edge crosses a land--water boundary. All specifications are
row-normalised so that $\sum_{j \in \mathcal{N}(i)} \tilde W_{ij} = 1$.
The four specifications are:

1. **Queen**: $\tilde W_{ij} = 1/n_i$ for all $j \in \mathcal{N}(i)$
2. **Rook**: restricts to horizontal/vertical neighbours only, $\tilde W_{ij} = 1/n_i^{(\text{rook})}$
3. **LC-eps**: $\tilde W_{ij} \propto (1 - \ell_{ij}) + \varepsilon\,\ell_{ij}$,\quad $\varepsilon = e^{-\alpha}$,\quad $\alpha$ CV-tuned
4. **LC-alpha**: $\tilde W_{ij} \propto e^{-\alpha |w_i - w_j|} + \delta$,\quad $\delta = 10^{-6}$,\quad $\alpha$ CV-tuned

In both LC-aware specifications $\alpha = 0$ recovers the queen baseline.
Cross-validation selected moderate values of $\alpha$:
$\hat\alpha_\text{eps} = `r fmt3(gf(r$nb_lc_eps,"alpha_opt"))`$ and
$\hat\alpha_\text{alpha} = `r fmt3(gf(r$nb_lc_alpha,"alpha_opt"))`$.
Figure 3 shows that at these tuned values the edge weights are only
moderately attenuated across the land--water boundary — the LC-aware
specifications do reduce cross-boundary communication, but not drastically.
Even so, Tables 3 and 4 show that all four specifications give nearly
identical predictive performance for albedo and SIF respectively. The
takeaway is that CV selected a moderate landcover penalty, and even that
moderate penalty made no meaningful difference to the results.

```{r nb_visualization, fig.cap="Left: unnormalised edge weight as a function of water fraction difference $|w_i - w_j|$ for the two LC-aware specifications at their CV-tuned $\\alpha$ values, alongside the flat queen baseline. Both LC-aware curves are nearly flat at the tuned $\\alpha$, explaining why results are insensitive to neighbourhood specification. Right: dominant landcover in the target grid, showing the land--water boundary where LC-aware weights would differ from queen.", fig.height=3.5, fig.width=9}

alpha_eps   <- gf(r$nb_lc_eps,   "alpha_opt")
alpha_alpha <- gf(r$nb_lc_alpha, "alpha_opt")

d_seq <- seq(0, 1, by = 0.01)

eps_val <- exp(-alpha_eps)
w_eps   <- ifelse(d_seq < 0.5, 1 + eps_val, eps_val)

w_alpha_val <- exp(-alpha_alpha * d_seq) + 1e-6
w_queen     <- rep(1, length(d_seq))

df_w <- data.frame(
  d      = rep(d_seq, 3),
  weight = c(w_queen, w_eps, w_alpha_val),
  Spec   = rep(c(
    "Queen (baseline)",
    sprintf("LC-eps (alpha=%.2f)", alpha_eps),
    sprintf("LC-alpha (alpha=%.2f)", alpha_alpha)
  ), each = length(d_seq))
)
df_w$Spec <- factor(df_w$Spec, levels = unique(df_w$Spec))

p_weights <- ggplot(df_w, aes(x = d, y = weight, color = Spec,
                               linetype = Spec)) +
  geom_line(linewidth = 0.9) +
  geom_vline(xintercept = 0.5, linetype = "dotted", color = "grey50") +
  annotate("text", x = 0.52, y = max(df_w$weight) * 0.95,
           label = "Land/water\nthreshold", hjust = 0, size = 2.5,
           color = "grey50") +
  scale_color_manual(values = c("steelblue", "orange", "tomato"), name = NULL) +
  scale_linetype_manual(values = c("solid", "dashed", "dotdash"), name = NULL) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(x     = expression("|"*w[i] - w[j]*"|"~"(water fraction difference)"),
       y     = "Unnormalised edge weight",
       title = "Edge weight vs water fraction difference") +
  theme_minimal(base_size = 9) +
  theme(legend.position  = "bottom",
        panel.grid.minor = element_blank(),
        plot.title       = element_text(size = 10, hjust = 0.5))

tg <- goebel2026::target_grid
dc <- tg$dominant_class
tg$dom_class <- dplyr::case_when(
  dc %in% c(1, 8)       ~ "Built/Bare",
  dc %in% c(6, 7)       ~ "Water",
  dc %in% c(2, 3, 4, 5) ~ "Green",
  TRUE                  ~ NA_character_
)

p_lc <- ggplot(tg[!is.na(tg$dom_class), ]) +
  geom_sf(aes(fill = dom_class), color = NA) +
  scale_fill_manual(
    values   = c("Built/Bare" = "#c4281b",
                 "Water"      = "#429ae4",
                 "Green"      = "#397e48"),
    name     = "Landcover",
    na.value = "grey90") +
  labs(title = "Dominant landcover (target grid)") +
  theme_void(base_size = 9) +
  theme(legend.position = "right",
        plot.title      = element_text(size = 10, hjust = 0.5))

p_weights | p_lc
```

```{r nb_albedo}
nbs_alb <- list(r$water_rsr, r$nb_rook, r$nb_lc_eps, r$nb_lc_alpha)

alpha_eps_hat   <- fmt3(gf(r$nb_lc_eps,   "alpha_opt"))
alpha_alpha_hat <- fmt3(gf(r$nb_lc_alpha, "alpha_opt"))

tab_nb_alb <- data.frame(
  Neighbourhood = c(
    "Queen (baseline)",
    "Rook",
    paste0("LC-eps (alpha=", alpha_eps_hat, ")"),
    paste0("LC-alpha (alpha=", alpha_alpha_hat, ")")
  ),
  rho      = sapply(nbs_alb, get_rho),
  phi      = sapply(nbs_alb, gf, "phi"),
  RMSE     = sapply(nbs_alb, gf, "rmse"),
  R2       = sapply(nbs_alb, gf, "r2"),
  Coverage = sapply(nbs_alb, gf, "coverage_95_obs"),
  beta_w   = sapply(nbs_alb, gf, "beta_hat", idx = 2)
)

kable(tab_nb_alb,
  format    = "pipe",
  digits    = 4,
  caption   = "Albedo predictive performance across neighbourhood weight specifications. All models include the water proportion covariate and RSR constraint at rho=1.",
  col.names = c("Neighbourhood", "rho", "phi",
                "RMSE", "R2", "Coverage (95%)",
                "beta_water"))
```

```{r nb_sif}
nbs_sif <- list(r$sif_rsr_rho1, r$sif_nb_rook, r$sif_nb_lc_eps, r$sif_nb_lc_alpha)

alpha_sif_eps_hat   <- fmt3(gf(r$sif_nb_lc_eps,   "alpha_opt"))
alpha_sif_alpha_hat <- fmt3(gf(r$sif_nb_lc_alpha, "alpha_opt"))

tab_nb_sif <- data.frame(
  Neighbourhood = c(
    "Queen (baseline)",
    "Rook",
    paste0("LC-eps (alpha=", alpha_sif_eps_hat, ")"),
    paste0("LC-alpha (alpha=", alpha_sif_alpha_hat, ")")
  ),
  rho     = sapply(nbs_sif, get_rho),
  phi     = sapply(nbs_sif, gf, "phi"),
  beta_w  = sapply(nbs_sif, gf, "beta_hat", idx = 2),
  mean_SE = sapply(nbs_sif, function(x) mean(x$posterior_se, na.rm = TRUE))
)

kable(tab_nb_sif,
  format    = "pipe",
  digits    = 4,
  caption   = "SIF posterior summaries across neighbourhood weight specifications. All models include the water proportion covariate, RSR constraint, and R^-1 weighting.",
  col.names = c("Neighbourhood", "rho", "phi",
                "beta_water", "Mean SE"))
```

\newpage

# Restricted Spatial Regression {#sec:rsr}

Spatial confounding arises when both the covariate and the spatial random
effect can explain the same variation in the response, leading to unstable
covariate estimates. In our application, water proportion is the covariate
of interest, and the land--water boundary creates a sharp spatial contrast
that both the covariate and the SAR random effect can absorb. When the model
is fit without any correction for this, the estimated water coefficient is
positive (Table 5), which is implausible for two reasons: first, it is
inconsistent with the SIF literature, which finds that water bodies have
lower photosynthetic activity than vegetated land; and second, it contradicts
the relationship observed at the coarse sounding scale, where the coefficient
is approximately $-0.6$, suggesting water pixels have lower SIF than land
pixels. A better approach should recover a result consistent with this
coarse-scale relationship. Several approaches to handling the confounding
are possible, each with drawbacks:

\begin{itemize}
  \item \textbf{Zeroing out water pixels entirely} would prevent the covariate
    from fitting water, but OCO-2 soundings physically integrate over
    mixed land--water footprints, so ignoring water would misrepresent
    the forward model.
  \item \textbf{Removing soundings over water} throws away potentially
    useful observations and would leave portions of the target grid
    unconstrained.
  \item \textbf{Restricting neighbourhood communication across the
    land--water boundary} (Section 2) reduces spatial smoothing across
    the boundary but does not resolve the covariate--residual competition,
    and is sensitive to how the boundary is defined given that water
    proportion is continuous.
  \item \textbf{Zeroing the residual over water pixels while allowing
    a covariate fit} is conceptually close to RSR in spirit but harder
    to implement cleanly and introduces additional modelling decisions
    about which pixels to constrain.
\end{itemize}

We use restricted spatial regression (RSR), which orthogonalises the spatial
residual against the covariate space so that $\hat\beta$ is identified by
covariate variation alone. We note that RSR is not without controversy:
it can introduce bias in the covariate estimate when the true data-generating
process does not respect the orthogonality constraint (Hodges \& Reich, 2010).
In our setting we adopt it as a pragmatic design choice to prevent the
spatial random effect from absorbing the land--water contrast, rather than
as a general recommendation for spatial confounding problems.

Table 5 illustrates the effect of RSR for SIF. Without
any spatial random effect (OLS) or with an unconstrained spatial effect
(naive), the estimated water coefficient is
$\hat\beta_\text{water} = `r fmt4(gf(r$sif_naive,"beta_hat",idx=2))`$, which is
positive — implying that water pixels have higher SIF than land pixels, a
physically implausible result. Applying RSR reverses the sign:
$\hat\beta_\text{water} = `r fmt4(gf(r$sif,"beta_hat",idx=2))`$, consistent with
water having lower photosynthetic activity than vegetated land. This sign
reversal suggests that without RSR the spatial random effect absorbs much
of the land--water contrast, leaving the covariate to explain residual
variation that is spuriously positive, though we acknowledge this
interpretation depends on the assumption that the orthogonality constraint
is approximately appropriate.

For albedo, where the ground truth is known, RSR also improves $R^2$ and
coverage (Table 2 in the main paper), providing empirical support that the
constraint produces better-calibrated uncertainty estimates as well as more
physically interpretable coefficients in this application.

```{r rsr_sif}
rsrs <- list(r$sif_naive, r$sif_rsr_rho1, r$sif)

tab_rsr <- data.frame(
  Model    = c("Naive (no RSR)",
               "RSR, rho=1",
               "RSR, rho tuned (canonical)"),
  rho      = sapply(rsrs, get_rho),
  phi      = sapply(rsrs, gf, "phi"),
  beta_int = sapply(rsrs, gf, "beta_hat", idx = 1),
  beta_w   = sapply(rsrs, gf, "beta_hat", idx = 2),
  mean_SE  = sapply(rsrs, function(x) mean(x$posterior_se, na.rm = TRUE))
)

kable(tab_rsr,
  format    = "pipe",
  digits    = 4,
  caption   = "SIF posterior summaries with and without RSR. The naive model (no RSR) estimates a positive water coefficient, which is physically implausible. RSR corrects the sign and produces stable estimates across rho specifications.",
  col.names = c("Model", "rho", "phi",
                "beta_0", "beta_water",
                "Mean SE"))
```

\newpage

# Model Diagnostics {#sec:diagnostics}

We compute standardized prediction residuals for the canonical SIF fit:

$$r_i = \frac{y_i - \hat{y}_i}{\sqrt{\widehat{\text{Var}}(y_i - \hat{y}_i)}}$$

where $\hat{y}_i = A_i \hat\mu$ is the fitted value and the predictive
variance combines posterior uncertainty in the fitted value with
per-sounding observation noise:

$$\widehat{\text{Var}}(y_i - \hat{y}_i) = \hat\sigma^2_e \left(A_i K^{-1} A_i^\top + [R^{-1}]_{ii}^{-1}\right)$$


```{r residual_diagnostic, fig.cap="Spatial residual diagnostic for the canonical SIF fit. Panel A: standardized prediction residuals mapped at sounding centroids; no spatial clustering is expected under correct specification. Panel B: standardized residual versus predictive SD; a flat loess indicates no heteroskedastic misspecification. Panel C: empirical variogram of standardized residuals with fitted exponential model (red); the dashed line marks the expected sill of 1 under no autocorrelation. Panel D: Q-Q plot against $N(0,1)$.", fig.height=7, fig.width=9}

diag <- .pkg_get("results_diagnostic_residuals")

df_resid <- data.frame(
  x         = diag$coords$x,
  y         = diag$coords$y,
  std_resid = diag$std_resid,
  pred_sd   = diag$pred_sd
)

clamp <- function(x, lo = -3.5, hi = 3.5) pmax(pmin(x, hi), lo)

p_map <- ggplot(df_resid, aes(x = x, y = y, colour = clamp(std_resid))) +
  geom_point(size = 0.9, alpha = 0.75) +
  scale_colour_distiller(palette = "RdBu", limits = c(-3.5, 3.5),
                         name = "Std.\nresid.\n(±3.5)") +
  coord_equal() +
  labs(title = "A: Standardized residuals",
       x = "Easting (m)", y = "Northing (m)") +
  theme_bw(base_size = 9) +
  theme(axis.text = element_text(size = 7))

p_sd <- ggplot(df_resid, aes(x = pred_sd, y = std_resid)) +
  geom_point(alpha = 0.3, size = 0.6) +
  geom_hline(yintercept = c(-2, 0, 2),
             linetype = c("dashed", "solid", "dashed"),
             colour = "steelblue") +
  geom_smooth(method = "loess", se = TRUE,
              colour = "tomato", linewidth = 0.8) +
  labs(title = "B: Std. residual vs. predictive SD",
       x = "Predictive SD", y = "Std. residual") +
  theme_bw(base_size = 9)

p_vgm <- ggplot(diag$vgm_emp, aes(x = dist, y = gamma)) +
  geom_point(aes(size = np), colour = "grey30") +
  scale_size_continuous(name = "Pairs", range = c(1, 3)) +
  geom_hline(yintercept = 1, linetype = "dashed",
             colour = "steelblue", linewidth = 0.8) +
  labs(title = "C: Variogram of std. residuals",
       x = "Distance (m)", y = "Semivariance") +
  theme_bw(base_size = 9)

if (!is.null(diag$vgm_fit)) {
  max_dist  <- max(diag$vgm_emp$dist)
  vgm_line  <- gstat::variogramLine(diag$vgm_fit, maxdist = max_dist, n = 200L)
  p_vgm <- p_vgm +
    geom_line(data = vgm_line, aes(x = dist, y = gamma),
              colour = "tomato", linewidth = 0.8, inherit.aes = FALSE)
}

p_qq <- ggplot(df_resid, aes(sample = std_resid)) +
  stat_qq(alpha = 0.3, size = 0.6) +
  stat_qq_line(colour = "tomato", linewidth = 0.8) +
  labs(title = "D: Q-Q plot",
       x = "Theoretical quantiles", y = "Sample quantiles") +
  theme_bw(base_size = 9)

(p_map | p_sd) / (p_vgm | p_qq)
```

The figure above shows the four diagnostic panels. The standardized residuals have
mean `r fmt3(diag$resid_mean)` and standard deviation `r fmt3(diag$resid_sd)`
(both close to their expected values of 0 and 1 respectively), and
`r sprintf("%.1f%%", 100 * diag$pct_gt2)` exceed $\pm 2$ in absolute value
(expected 5% under normality). The Q-Q plot shows approximate normality with
mild heavy tails, consistent with the heterogeneous observation density across
soundings. Panel B shows no systematic relationship between residual magnitude
and predictive uncertainty, indicating no heteroskedastic misspecification.
The empirical variogram is flat across all lag distances with a sill near 1,
confirming no residual spatial autocorrelation. Together these diagnostics
support the model specification and confirm that the SAR prior adequately
absorbs the spatial structure in the data.

\newpage

# Forward Operator Sensitivity {#sec:forward_operator}

The aggregation matrix $A$ discretises the continuous change-of-support
integral by replacing each intersection integral with a weighted average
over the intersecting latent cell. The default specification uses uniform
weights within each intersection ($g_i \equiv 1$). Here we assess sensitivity
to this choice by considering a family of super-Gaussian sensor response
functions parameterised by a width $\tau$.

**Sensor response function.** For sounding $i$ with centroid $c_i$ and
affine map $M_i^{-1}$ transforming footprint coordinates to a canonical
unit square, the response function is:
$$g_i(s) = \exp\!\left(-\frac{\|M_i^{-1}(s - c_i)\|^2}{2\tau^2}\right)$$
As $\tau \to \infty$ the response approaches uniform ($g_i \equiv 1$,
the default). As $\tau \to 0$ the response concentrates at the footprint
centre, approaching the centroid approximation in the limit.
The aggregation weight for cell $j$ under sounding $i$ is then
$a_{ij} \propto \int_{D_i^{(o)} \cap D_j^{(\ell)}} g_i(s)\,ds$,
normalised so that $\sum_j a_{ij} = 1$. Each intersection integral is
evaluated numerically by triangulating the intersection polygon and
applying quasi-Monte Carlo quadrature within each triangle.

Figure 4 shows $g_i$ evaluated on a representative rectangular
OCO-2 footprint for each $\tau$ value considered. The five tau values tested
span from nearly uniform ($\tau=0.5$) to sharply peaked ($\tau=0.01$).

```{r gA_viz, fig.cap="Super-Gaussian sensor response $g_i(s)$ evaluated on a representative OCO-2 footprint for each $\\tau$ value. Uniform ($\\tau = \\infty$, top left) assigns equal weight everywhere; smaller $\\tau$ concentrates weight at the footprint centre. The centroid approximation (bottom right) is the limiting case.", fig.height=4, fig.width=9}

nx <- 60; ny <- 40
xs <- seq(-1, 1, length.out = nx)
ys <- seq(-1, 1, length.out = ny)
grid_uv <- expand.grid(u = xs, v = ys)

tau_vals_viz <- c(Inf, 0.5, 1/3, 0.2, 0.1, 0.01)
tau_labs     <- c("Uniform (tau=inf)", "tau=0.5", "tau=0.33",
                  "tau=0.2",           "tau=0.1", "tau=0.01 (centroid)")

df_g <- do.call(rbind, lapply(seq_along(tau_vals_viz), function(k) {
  tau <- tau_vals_viz[k]
  g   <- if (is.infinite(tau)) rep(1, nrow(grid_uv)) else
           exp(-(grid_uv$u^2 + grid_uv$v^2) / (2 * tau^2))
  data.frame(u     = grid_uv$u,
             v     = grid_uv$v,
             g     = g / max(g),
             label = tau_labs[k])
}))
df_g$label <- factor(df_g$label, levels = tau_labs)

ggplot(df_g, aes(x = u, y = v, fill = g)) +
  geom_tile() +
  facet_wrap(~ label, nrow = 2) +
  scale_fill_gradient(low = "white", high = "tomato",
                      name = expression(g[i](s)~"(normalised)")) +
  coord_equal() +
  labs(x = "Footprint u-coordinate",
       y = "Footprint v-coordinate") +
  theme_minimal(base_size = 9) +
  theme(legend.position  = "bottom",
        panel.grid       = element_blank(),
        strip.text       = element_text(size = 8))
```

## Albedo validation

Because the albedo data has a known generating field, we can directly assess
the impact of forward operator misspecification on RMSE and coverage.
Figure 5 and Table 6 show results for
albedo as $\tau$ decreases from 0.5 (broad response, close to uniform) to
the centroid limit. Coverage degrades monotonically from
`r pct(gf(r$water_rsr_cv,"coverage_95_obs"))` at the uniform baseline to
`r pct(gf(r$alb_gA_cent,"coverage_95_obs"))` at the centroid limit, while
RMSE increases by
`r fmt4(gf(r$alb_gA_cent,"rmse") - gf(r$water_rsr_cv,"rmse"))` absolute
units. This confirms that forward operator misspecification primarily damages
uncertainty quantification rather than point prediction accuracy.

```{r gA_albedo, fig.cap="Coverage (95\\%) and RMSE as functions of the sensor response width $\\tau$ for the albedo validation experiment. The uniform baseline ($\\tau = \\infty$, leftmost point) achieves near-nominal coverage; coverage degrades monotonically as $\\tau \\to 0$ (centroid limit, rightmost point).", fig.height=3.5, fig.width=7}

tau_vals  <- c(Inf, 0.5, 1/3, 0.2, 0.1, 0)
cov_vals  <- c(gf(r$water_rsr_cv, "coverage_95_obs"),
               gf(r$alb_gA_tau05,  "coverage_95_obs"),
               gf(r$alb_gA_tau033, "coverage_95_obs"),
               gf(r$alb_gA_tau02,  "coverage_95_obs"),
               gf(r$alb_gA_tau01,  "coverage_95_obs"),
               gf(r$alb_gA_cent,   "coverage_95_obs"))
rmse_vals <- c(gf(r$water_rsr_cv, "rmse"),
               gf(r$alb_gA_tau05,  "rmse"),
               gf(r$alb_gA_tau033, "rmse"),
               gf(r$alb_gA_tau02,  "rmse"),
               gf(r$alb_gA_tau01,  "rmse"),
               gf(r$alb_gA_cent,   "rmse"))

tau_labels <- c("Uniform\n(baseline)", "0.5", "0.33", "0.2", "0.1", "Centroid")
df_gA <- data.frame(
  tau      = factor(tau_labels, levels = tau_labels),
  coverage = cov_vals,
  rmse     = rmse_vals
)

p_cov <- ggplot(df_gA, aes(x = tau, y = coverage, group = 1)) +
  geom_hline(yintercept = 0.95, linetype = "dashed", color = "gray50") +
  geom_line(color = "steelblue", linewidth = 0.9) +
  geom_point(color = "steelblue", size = 2.5) +
  scale_y_continuous(limits = c(0.5, 1.0),
                     labels = scales::percent_format(accuracy = 1)) +
  labs(x = expression(tau), y = "Coverage (95%)") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank())

p_rmse <- ggplot(df_gA, aes(x = tau, y = rmse, group = 1)) +
  geom_line(color = "tomato", linewidth = 0.9) +
  geom_point(color = "tomato", size = 2.5) +
  labs(x = expression(tau), y = "RMSE") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank())

p_cov | p_rmse
```

```{r gA_albedo_table}
gAs_alb <- list(r$water_rsr_cv, r$alb_gA_tau05, r$alb_gA_tau033,
                r$alb_gA_tau02, r$alb_gA_tau01, r$alb_gA_cent)

tab_gA_alb <- data.frame(
  tau      = c("Uniform (baseline)", "0.5", "0.33", "0.2", "0.1", "Centroid"),
  rho      = sapply(gAs_alb, get_rho),
  phi      = sapply(gAs_alb, gf, "phi"),
  RMSE     = rmse_vals,
  R2       = sapply(gAs_alb, gf, "r2"),
  Coverage = cov_vals
)

kable(tab_gA_alb,
  format    = "pipe",
  digits    = 4,
  caption   = "Albedo predictive performance under forward operator misspecification. tau controls the width of the super-Gaussian sensor response function; tau -> 0 corresponds to the centroid approximation. Coverage degrades monotonically while RMSE increases modestly.",
  col.names = c("tau", "rho", "phi",
                "RMSE", "R2", "Coverage (95%)"))
```

\newpage

## SIF sensitivity

For SIF, ground truth is unavailable, so we assess sensitivity through
changes in posterior mean and uncertainty. The table below shows
that as $\tau$ decreases, $\hat\rho$ increases toward 1 and $\hat\phi$
decreases, reflecting the model compensating for a more peaked forward
operator by increasing spatial smoothness. The water coefficient
$\hat\beta_\text{water}$ remains stable across all specifications
(range: `r fmt4(min(sapply(list(r$sif, r$sif_gA_tau05, r$sif_gA_tau033, r$sif_gA_tau02, r$sif_gA_tau01, r$sif_gA_tau001), gf, "beta_hat", idx=2)))`
to
`r fmt4(max(sapply(list(r$sif, r$sif_gA_tau05, r$sif_gA_tau033, r$sif_gA_tau02, r$sif_gA_tau01, r$sif_gA_tau001), gf, "beta_hat", idx=2)))`),
suggesting that the posterior mean is robust to the exact form of the sensor
response function even as uncertainty estimates vary.

```{r gA_sif}
gAs_sif <- list(r$sif, r$sif_gA_tau05, r$sif_gA_tau033,
                r$sif_gA_tau02, r$sif_gA_tau01, r$sif_gA_tau001)

tab_gA_sif <- data.frame(
  tau     = c("Uniform (baseline)", "0.5", "0.33", "0.2", "0.1", "0.01"),
  rho     = sapply(gAs_sif, get_rho),
  phi     = sapply(gAs_sif, gf, "phi"),
  beta_w  = sapply(gAs_sif, gf, "beta_hat", idx = 2),
  mean_SE = sapply(gAs_sif, function(x) mean(x$posterior_se, na.rm = TRUE))
)

kable(tab_gA_sif,
  format    = "pipe",
  digits    = 4,
  caption   = "SIF posterior summaries under forward operator misspecification. The water coefficient is stable across all specifications; mean posterior SE decreases as tau -> 0 reflecting increased smoothing.",
  col.names = c("tau", "rho", "phi",
                "beta_water", "Mean SE"))
```

## Kriging comparison

**Kriging setup.** We fit universal kriging with the water proportion as a
covariate, using the model $y(s) = \beta_0 + \beta_1 w(s) + \varepsilon(s)$
where $\varepsilon$ is a zero-mean stationary Gaussian process. Observations
are treated as **point support at sounding centroids** — the standard
geostatistical assumption that ignores the areal nature of the footprints.
An exponential variogram model is fitted empirically to the sounding-level
residuals (cutoff 15 km, bin width 500 m), and universal kriging predictions
are made at the 330 m target grid cell centroids.

The fitted variogram parameters for albedo are nugget =
`r fmt4(r$krig_alb$vgm_fit$psill[1])`,
partial sill = `r fmt4(r$krig_alb$vgm_fit$psill[2])`,
range = `r fmt1(r$krig_alb$vgm_fit$range[2])` m.

```{r kriging_variogram, fig.cap="Empirical variogram (points) and fitted exponential model (line) for the semi-synthetic albedo data. The nugget reflects measurement error; the range indicates the spatial scale of correlation.", fig.height=3, fig.width=5}

vgm_emp <- r$krig_alb$vgm_emp
vgm_fit <- r$krig_alb$vgm_fit

lag_seq  <- seq(0, max(vgm_emp$dist) * 1.05, length.out = 200)
nugget   <- vgm_fit$psill[1]
psill    <- vgm_fit$psill[2]
range_   <- vgm_fit$range[2]
vgm_line <- data.frame(
  dist  = lag_seq,
  gamma = nugget + psill * (1 - exp(-lag_seq / range_))
)

ggplot() +
  geom_line(data  = vgm_line,
            aes(x = dist / 1000, y = gamma),
            color = "steelblue", linewidth = 0.9) +
  geom_point(data  = vgm_emp,
             aes(x = dist / 1000, y = gamma, size = np),
             color = "grey30", alpha = 0.7) +
  scale_size_continuous(name = "n pairs", range = c(1, 4)) +
  labs(x     = "Lag distance (km)",
       y     = expression(hat(gamma)(h)~"(semivariance)"),
       title = "Albedo empirical variogram") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank(),
        legend.position  = "right")
```

The Bayesian approach achieves substantially better coverage
(`r pct(gf(r$water_rsr_cv,"coverage_95_obs"))` vs.\
`r pct(gf(r$krig_alb,"coverage_95_obs"))`) with similar RMSE. The coverage
gap arises because kriging treats each observation as a point, so its
prediction variance reflects interpolation uncertainty but not the
change-of-support uncertainty from aggregating an areal field — leading to
intervals that are systematically too narrow (Goovaerts, 2008).

```{r kriging_table}
krigs <- list(r$water_rsr_cv, r$alb_gA_cent, r$krig_alb)

tab_krig <- data.frame(
  Method   = c("Bayesian (uniform A, baseline)",
               "Bayesian (centroid A)",
               "Universal kriging"),
  RMSE     = sapply(krigs, gf, "rmse"),
  R2       = sapply(krigs, gf, "r2"),
  Coverage = sapply(krigs, gf, "coverage_95_obs")
)

kable(tab_krig,
  format    = "pipe",
  digits    = 4,
  caption   = "Comparison of Bayesian downscaling and universal kriging on the albedo validation experiment. The centroid approximation row shows performance when the aggregation matrix uses footprint centroids rather than intersection areas. Kriging coverage is substantially below nominal; the Bayesian approach with correct change-of-support aggregation achieves near-nominal coverage.",
  col.names = c("Method", "RMSE", "R2", "Coverage (95%)"))
```

## Forward operator misspecification experiment

To assess whether the choice of $A$ matters in practice, we conduct a
controlled 2x2 experiment. We generate two sets of synthetic
observations from the known albedo field: one using the uniform aggregation
matrix ($A_\text{unif}$) and one using the $g$-weighted matrix with
$\tau = 1/3$ ($A_g$). We then fit the model under each $A$ specification
to each dataset, giving four combinations. If forward operator
misspecification is consequential, the off-diagonal cells (misspecified
fitting $A$) should perform worse than the diagonal cells (correctly
specified).

The table below shows RMSE and $R^2$ for all four combinations.
The diagonal entries outperform the corresponding off-diagonal entries in
both cases, confirming that misspecification is detectable. The effect is
asymmetric: generating with $A_g$ and fitting with $A_\text{unif}$ incurs a
larger RMSE penalty than the reverse, suggesting that over-smoothing the
forward operator (uniform $A$ applied to data generated with a peaked $g$)
is more harmful than under-smoothing. However, all four cells achieve $R^2 >
0.70$, so the absolute magnitude of the effect is modest relative to the
overall predictive performance — consistent with the tau sensitivity results
above.

```{r contingency_table}
ct_uu <- r$ct_uu
ct_ug <- r$ct_ug
ct_gu <- r$ct_gu
ct_gg <- r$ct_gg

tab_ct <- data.frame(
  gen  = c("Uniform", "Uniform", "g-weighted", "g-weighted"),
  fit  = c("Uniform", "g-weighted", "Uniform", "g-weighted"),
  rmse = c(gf(ct_uu,"rmse"), gf(ct_ug,"rmse"),
           gf(ct_gu,"rmse"), gf(ct_gg,"rmse")),
  r2   = c(gf(ct_uu,"r2"),   gf(ct_ug,"r2"),
           gf(ct_gu,"r2"),   gf(ct_gg,"r2")),
  phi  = c(gf(ct_uu,"phi"),  gf(ct_ug,"phi"),
           gf(ct_gu,"phi"),  gf(ct_gg,"phi"))
)

kable(tab_ct,
  format    = "pipe",
  digits    = 4,
  caption   = "2x2 misspecification experiment: RMSE and R2 for each combination of data-generating A (rows) and fitting A (columns). Diagonal entries are correctly specified; off-diagonal entries are misspecified. The diagonal outperforms in both cases, with the gen=g/fit=uniform cell showing the largest penalty.",
  col.names = c("Generating A", "Fitting A", "RMSE", "R2", "phi"))
```

```{r contingency_maps, fig.cap="Posterior mean albedo maps for the 2x2 misspecification experiment. Rows correspond to the data-generating A (uniform, top; g-weighted, bottom); columns to the fitting A (uniform, left; g-weighted, right). Diagonal panels are correctly specified. Spatial patterns are broadly similar across all four cells; differences are most visible at footprint boundaries in the off-diagonal panels.", fig.height=6, fig.width=9}

covered <- ct_uu$covered
tg_sf   <- goebel2026::target_grid[covered, ]

# Shared colour scale across all four panels
ct_shared_lims <- range(
  c(ct_uu$posterior_mean[covered], ct_ug$posterior_mean[covered],
    ct_gu$posterior_mean[covered], ct_gg$posterior_mean[covered]),
  na.rm = TRUE
)

make_ct_map <- function(ct, title_str) {
  tg_sf$val <- ct$posterior_mean[covered]
  ggplot(tg_sf) +
    geom_sf(aes(fill = val), color = NA) +
    scale_fill_viridis_c(name = "Albedo",
                         limits = ct_shared_lims,
                         option = "magma") +
    labs(title    = title_str,
         subtitle = sprintf("RMSE=%.4f  R2=%.3f", ct$rmse, ct$r2)) +
    theme_void(base_size = 9) +
    theme(legend.position = "right")
}

p_uu <- make_ct_map(ct_uu, "Gen: Uniform    | Fit: Uniform")
p_ug <- make_ct_map(ct_ug, "Gen: Uniform    | Fit: g-weighted")
p_gu <- make_ct_map(ct_gu, "Gen: g-weighted | Fit: Uniform")
p_gg <- make_ct_map(ct_gg, "Gen: g-weighted | Fit: g-weighted")

(p_uu | p_ug) / (p_gu | p_gg)
```

\newpage

# Rho Sensitivity {#sec:rho}

The SAR spatial range parameter $\rho$ controls the decay of spatial
correlation: $\rho = 1$ corresponds to an intrinsic (improper) prior with
no decay, while smaller values introduce exponential decay in correlations
with distance. In the main analysis we tune $\rho$ jointly with $\phi$ via
cross-validation.

The table below shows albedo results at $\rho = 1$ (intrinsic)
and at the CV-tuned value $\hat\rho =
`r fmt3(get_rho(r$water_rsr_cv))`$. The differences in RMSE, $R^2$, and
coverage are small (less than 1 percentage point in coverage), confirming
that the intrinsic prior is a reasonable default. We nevertheless prefer
the tuned $\rho$ in the main analysis as it provides marginally better
coverage calibration. For SIF, the canonical model uses
$\hat\rho = `r fmt3(get_rho(r$sif))`$; fixing $\rho = 0.95$ gives
essentially identical posterior means and $\hat\beta_\text{water}$ values
(Table 8 in the main results summary).

```{r rho_sensitivity}
rhos <- list(r$water_rsr, r$water_rsr_cv, r$water, r$water_rho_cv)

tab_rho <- data.frame(
  Model    = c("Water + RSR, rho=1",
               "Water + RSR, rho tuned",
               "Water (no RSR), rho=1",
               "Water (no RSR), rho tuned"),
  rho      = sapply(rhos, get_rho),
  phi      = sapply(rhos, gf, "phi"),
  RMSE     = sapply(rhos, gf, "rmse"),
  R2       = sapply(rhos, gf, "r2"),
  Coverage = sapply(rhos, gf, "coverage_95_obs")
)

kable(tab_rho,
  format    = "pipe",
  digits    = 4,
  caption   = "Albedo predictive performance at fixed rho=1 versus CV-tuned rho, with and without RSR. Differences are small in all metrics, confirming that the intrinsic prior (rho=1) is a reasonable default.",
  col.names = c("Model", "rho", "phi",
                "RMSE", "R2", "Coverage (95%)"))
```
