---
title: "Supplement: Robustness Checks and Model Extensions"
output:
  pdf_document:
    toc: true
    number_sections: true
    keep_tex: false
header-includes:
  - \usepackage{booktabs}
  - \usepackage{amsmath}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 5,
  fig.height = 3.5
)

library(goebel2026)
library(ggplot2)
library(knitr)
library(sf)

# ------------------------------------------------------------------
# Shared objects used across all sections
# ------------------------------------------------------------------

`%||%` <- function(x, y) if (!is.null(x)) x else y

load_result <- function(nm) {
  env <- new.env(parent = emptyenv())
  data(list = nm, package = "goebel2026", envir = env)
  env[[nm]]
}

perf_row <- function(res, label = NULL) {
  data.frame(
    Model    = label %||% res$run_name,
    rho      = round(res$rho_opt          %||% NA_real_, 4),
    phi      = round(res$phi              %||% NA_real_, 4),
    RMSE     = round(res$rmse             %||% NA_real_, 5),
    R2       = round(res$r2               %||% NA_real_, 4),
    Coverage = round(res$coverage_95_obs  %||% NA_real_, 4),
    stringsAsFactors = FALSE
  )
}
```

\newpage

# Spatially Blocked Cross-Validation {#sec-blocking}

## Motivation

The baseline model selects the smoothing parameter $\hat\phi$ by 10-fold
cross-validation, assigning soundings to folds by simple random sampling.
Because nearby soundings overlap many of the same fine-resolution pixels,
a randomly held-out sounding is partially explained by spatially proximate
soundings that remain in the training set. This leakage can cause CV to
underestimate generalisation error and favour overly rough solutions.

Spatially blocked CV addresses this by partitioning soundings into $K$
geographically contiguous blocks before assigning fold labels, so that
all soundings in a held-out fold are spatially separated from the training
soundings. The blocks are defined by $k$-means clustering of sounding
centroids (stored in \texttt{setup\_albedo\$blocked\_folds}).

## Fold structure

Figure \ref{fig:fold-map} shows the spatial arrangement of the 10 blocked
folds. Each sounding polygon is coloured by its fold assignment; the target
grid is shown in grey as a reference for the domain extent.

```{r fold-map, fig.cap="Spatial fold assignments for the blocked cross-validation scheme. Each sounding polygon is coloured by fold; the grey background shows the target grid domain.", fig.height=4.5}

soundings_sf      <- goebel2026::setup_shared$soundings_proj
blocked_folds     <- goebel2026::setup_albedo$blocked_folds
soundings_sf$fold <- factor(blocked_folds)

ggplot() +
  geom_sf(data = goebel2026::target_grid, fill = "grey92", colour = NA) +
  geom_sf(data = soundings_sf, aes(fill = fold), colour = NA, alpha = 0.8) +
  scale_fill_brewer(palette = "Paired", name = "Fold") +
  theme_minimal(base_size = 10) +
  theme(axis.text  = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank())
```

## Performance comparison

```{r blocking-table}
res_queen   <- load_result("results_water_rsr_rho1")
res_blocked <- load_result("results_cv_blocked")

tab <- rbind(
  perf_row(res_queen,   "Random CV (baseline)"),
  perf_row(res_blocked, "Blocked CV")
)
rownames(tab) <- NULL

knitr::kable(
  tab, booktabs = TRUE,
  col.names = c("Model", "$\\hat\\rho$", "$\\hat\\phi$",
                "RMSE", "$R^2$", "Coverage (95\\%)"),
  escape  = FALSE,
  caption = "Random vs.\\ spatially blocked CV. Performance computed
             over pixels with at least one overlapping sounding."
)
```

Blocked CV selects a substantially smaller $\hat\phi$ than random CV
(`r round(res_blocked$phi, 1)` vs.\ `r round(res_queen$phi, 1)`),
consistent with the expected direction: withholding the overlap-rich
central strip forces the CV objective to favour less smoothing, since
held-out predictions must rely more heavily on spatial extrapolation.
Despite this, downstream predictive performance (RMSE, $R^2$, coverage)
is nearly identical between the two schemes. This robustness is
attributable to the footprint-averaging structure of the forward
operator: each sounding integrates over many fine-resolution pixels,
so spatial leakage between randomly assigned folds is already limited.
We therefore use random CV in the main analysis as the simpler and
computationally cheaper option.

\newpage

# Soft Penalty on Spatial Confounding {#sec-softrsr}

## Hard vs.\ soft RSR

Hard restricted spatial regression (RSR) enforces the orthogonality
constraint
$$
X^{(o)\top} A\, r^{(\ell)} = 0
$$
exactly via a Schur-complement projection of the posterior. Every posterior
draw satisfies this constraint with probability one. While this cleanly
removes covariate signal from the spatial residual, it is an all-or-nothing
operation: there is no mechanism to allow partial confounding when the data
weakly support it.

The soft-penalty variant replaces the hard projection with a quadratic
penalty added directly to the prior precision,
$$
\Lambda_{\tau,\alpha}
  = \tau\,(I - W)^\top(I - W)
  + \alpha\,\tilde{C}^\top\tilde{C},
  \qquad
  \tilde{C} := \frac{X^{(o)\top}A}{\|X^{(o)\top}A\|_F},
$$
where $\alpha \geq 0$ controls the strength of the confounding penalty and
$\tilde{C}$ is the Frobenius-normalised constraint matrix. At $\alpha = 0$
this recovers the unconstrained SAR prior; as $\alpha \to \infty$ the
penalty dominates and the posterior approaches the hard-RSR solution. The
penalty adds only a rank-$q$ correction to the precision (where $q$ is the
number of covariates, here $q = 2$), so the sparsity structure of the
spatial precision is preserved.

Both $\phi$ and $\alpha$ are selected jointly by a grid search over
$\rho \in \{0.70, 0.90, 0.95, 0.99\}$ and
$\log\alpha \in \{-4, -3, \ldots, 4\}$, with $\phi$ profiled out at each
grid point via CV.

## CV surface

```{r softrsr-surface, fig.cap="Cross-validation MSE surface over the soft-penalty grid. Fill colour shows CV MSE at each $(\\rho, \\log\\alpha)$ point with $\\phi$ profiled out. The white star marks the selected optimum.", fig.width=5.5, fig.height=3.2}

res_soft <- load_result("results_softrsr_albedo")

grid_df   <- res_soft$grid
best_i    <- which.min(grid_df$cv_mse)

ggplot(grid_df, aes(x = log_alpha, y = factor(rho), fill = cv_mse)) +
  geom_tile(colour = "white", linewidth = 0.3) +
  geom_point(data = grid_df[best_i, ],
             aes(x = log_alpha, y = factor(rho)),
             shape = 8, colour = "white", size = 3, inherit.aes = FALSE) +
  scale_fill_viridis_c(option = "magma", direction = -1, name = "CV MSE") +
  scale_x_continuous(breaks = seq(-4, 4, by = 2)) +
  labs(x = expression(log~alpha), y = expression(rho)) +
  theme_minimal(base_size = 10) +
  theme(panel.grid = element_blank())
```

Figure \ref{fig:softrsr-surface} shows the CV MSE across the grid. The
surface is relatively flat across a wide range of $\alpha$ for each $\rho$,
with the minimum at
$\rho = `r res_soft$rho_opt`$,
$\alpha = `r round(res_soft$alpha_opt, 3)`$.
The insensitivity to $\alpha$ suggests the data do not strongly distinguish
between partial and full confounding adjustment in this setting.

## Performance comparison

```{r softrsr-table}
res_hard <- load_result("results_water_rsr_rho1")

tab <- rbind(
  perf_row(res_hard, "Hard RSR (baseline)"),
  perf_row(res_soft, "Soft penalty RSR")
)
rownames(tab) <- NULL

knitr::kable(
  tab, booktabs = TRUE,
  col.names = c("Model", "$\\hat\\rho$", "$\\hat\\phi$",
                "RMSE", "$R^2$", "Coverage (95\\%)"),
  escape  = FALSE,
  caption = "Hard RSR vs.\\ soft penalty. The soft model tunes $\\alpha$
             jointly with $\\phi$ via grid search CV."
)
```

\newpage

# Landcover-Aware Neighbor Weights {#sec-neighbors}

## Motivation

The standard SAR prior uses queen adjacency with uniform row-normalised
weights: $\widetilde W_{ij} = 1/n_i$ for all queen-adjacent pixel pairs,
regardless of land-cover type. One concern is that this couples land and
water pixels symmetrically, pulling water-pixel estimates toward their
land neighbors. Once water proportion is included as a covariate the
model already has a direct mechanism to separate land and water signal,
but the prior still shares spatial strength across boundaries.

We evaluate three modifications to the neighbor structure, all using
water covariate + RSR + $\rho = 1$ as the base model:

- **Rook adjacency**: restricts neighbors to the four cardinal directions
  (no diagonals), reducing each interior pixel from 8 to 4 neighbors.

- **Epsilon-floor weights** (variant A): within-type edges receive weight
  proportional to 1 and cross-type edges proportional to
  $\varepsilon = e^{-\alpha}$, with $\alpha \geq 0$ tuned by CV.
  At $\alpha = 0$ this recovers uniform queen weights; large $\alpha$
  drives cross-boundary weights toward zero.

- **Soft-alpha weights** (variant B): edge weights decay continuously with
  the difference in water fraction,
  $w_{ij} \propto \exp(-\alpha\,|f_i - f_j|)$,
  where $f_i$ is the water fraction of pixel $i$ and $\alpha$ is CV-tuned.
  This variant uses the full continuous water fraction rather than a binary
  land/water indicator.

## Weight structure

```{r weight-illustration, fig.cap="Edge weight as a function of neighbor water fraction $f_j$, for a focal pixel with water fraction $f_i = 0.8$. The dotted vertical line marks the land/water threshold at 0.5. Alpha values shown are CV-selected optima from the package results.", fig.width=5, fig.height=3}

res_eps   <- load_result("results_neighbor_lc_eps")
res_alpha <- load_result("results_neighbor_lc_alpha")

alpha_eps   <- res_eps$tags$alpha   %||% 2
alpha_alpha <- res_alpha$tags$alpha %||% 2

f_i <- 0.8
f_j <- seq(0, 1, length.out = 200)

is_water_i <- f_i > 0.5
eps  <- exp(-alpha_eps)
w_A  <- ifelse((f_j > 0.5) == is_water_i, 1, eps)
w_A  <- w_A / mean(w_A)

w_B  <- exp(-alpha_alpha * abs(f_j - f_i)) + 1e-6
w_B  <- w_B / mean(w_B)

df_ill <- data.frame(
  f_j    = rep(f_j, 3),
  weight = c(rep(1, length(f_j)), w_A, w_B),
  model  = rep(c("Queen (uniform)",
                 sprintf("Epsilon-floor (alpha=%.2f)", alpha_eps),
                 sprintf("Soft-alpha (alpha=%.2f)",    alpha_alpha)),
               each = length(f_j))
)

ggplot(df_ill, aes(x = f_j, y = weight, colour = model, linetype = model)) +
  geom_line(linewidth = 0.8) +
  geom_vline(xintercept = 0.5, linetype = "dotted", colour = "grey50") +
  scale_colour_manual(values = c("grey50", "#E69F00", "#0072B2")) +
  scale_linetype_manual(values = c("dashed", "solid", "solid")) +
  labs(x = expression("Neighbor water fraction " * f[j]),
       y = "Relative edge weight",
       colour = NULL, linetype = NULL) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 8))
```

## Performance comparison

```{r neighbor-table}
res_queen <- load_result("results_water_rsr_rho1")
res_rook  <- load_result("results_neighbor_rook")
res_lc    <- load_result("results_neighbor_lc")

tab <- rbind(
  perf_row(res_queen, "Queen, uniform (baseline)"),
  perf_row(res_rook,  "Rook, uniform"),
  perf_row(res_lc,    "Queen, landcover binary"),
  perf_row(res_eps,   sprintf("Queen, epsilon-floor (alpha=%.2f)", alpha_eps)),
  perf_row(res_alpha, sprintf("Queen, soft-alpha (alpha=%.2f)",    alpha_alpha))
)
rownames(tab) <- NULL

knitr::kable(
  tab, booktabs = TRUE,
  col.names = c("Model", "$\\hat\\rho$", "$\\hat\\phi$",
                "RMSE", "$R^2$", "Coverage (95\\%)"),
  escape  = FALSE,
  caption = "Neighbor structure sensitivity. All models use water covariate
             + RSR + $\\rho = 1$. Alpha values shown are CV-selected optima."
)
```

All landcover-aware variants perform comparably to the uniform queen
baseline. The tuned $\alpha$ values indicate a mild preference for
down-weighting cross-boundary edges, but the effect on RMSE and $R^2$
is negligible. This is consistent with the water covariate already
capturing the dominant land--water contrast in the latent field, leaving
little residual structure for the weight matrix to exploit.
